---
title: "Session 9.2: Spatio-temporal modelling for area data"
params: 
   conference: "Bayesian modelling for Spatial and Spatio-temporal data"
   location: "Imperial College London"
   date: ""
   short_title: "MSc Epidemiology"
output:
   xaringan::moon_reader: 
    includes: 
       # This line adds a logo based on the format selected in the file 'assets/include_logo.html'
       in_header: "assets/latex_macros.html" 
       # NB: the actual options (eg placement of the logo and actual logo file) can be changed there
     #  after_body: "assets/insert-logo.html" # Monica commented this
    seal: false
    yolo: no
    lib_dir: libs
    nature:
      beforeInit: ["assets/remark-zoom.js","https://platform.twitter.com/widgets.js"]
      navigation:
        scroll: false # disable slide transitions by scrolling
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
      titleSlideClass:
      - center
      - middle
    self_contained: false 
    css:
    - "assets/beamer.css"
editor_options: 
  chunk_output_type: console
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 999)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
#options(htmltools.preserve.raw = FALSE)
# https://stackoverflow.com/questions/65766516/xaringan-presentation-not-displaying-html-widgets-even-when-knitting-provided-t
```

```{r echo=F, message=FALSE, warning=FALSE, comment=NA}
source("assets/setup.R")
library(INLA)
xaringanExtra::use_panelset()
bibfile=RefManageR::ReadBib("~/Dropbox/TEACHING/YEAR_2023/Bayes_Spatial_2023/Material/Session9.2/Biblio.bib",check = FALSE)
```

class: title-slide

# `r rmarkdown::metadata$title``r vspace("10px")` `r rmarkdown::metadata$subtitle`

## `r rmarkdown::metadata$author`

### `r rmarkdown::metadata$institute`    

### `r rmarkdown::metadata$params$conference`, `r rmarkdown::metadata$params$location` 

<!-- Can also separate the various components of the extra argument 'params', eg as in 
### `r paste(rmarkdown::metadata$params, collapse=", ")`
-->

`r ifelse(is.null(rmarkdown::metadata$params$date),format(Sys.Date(),"%e %B %Y"),rmarkdown::metadata$params$date)`

---

layout: true  

.my-footer[ 
.alignleft[ 
&nbsp; &copy; Marta Blangiardo | Monica Pirani 
]
.aligncenter[
`r rmarkdown::metadata$params$short_title` 
]
.alignright[
`r rmarkdown::metadata$params$conference`, `r short_date` 
]
] 

```{css,echo=FALSE, eval=FALSE}
.red {
  color: red;
}
.blue {
  color: 0.14 0.34 0.55;
}

.content-box-blue { background-color: #F0F8FF; }

}
```

---

# Learning Objectives

At the end of this session you should be able to:

`r vspace("10px")`


- Explain how to extend spatial or temporal models to spatio-temporal models 


`r vspace("10px")`


- Fit Bayesian space-time models using `R-INLA`


`r vspace("20px")`


The topics covered in this lecture can be found in Chapter 7 of the book Spatial and **Spatio-Temporal Bayesian Models with `R-INLA`**.  

---

# Outline 

`r vspace("30px")`

1\. [Temporal dependence](#temporal-structure)

`r vspace("30px")`

2\. [From space to space-time](#space-to-time)

`r vspace("30px")`

3\. [Type of interactions](#interactions)

---

name: temporal-structure
  
`r vspace("250px")`

.myblue[.center[.huge[
**Temporal dependence**]]]

---

# Temporal dependence

- Similarly to spatial dependence, it is sometimes necessary to model temporal dependence of data or of parameters:

`r vspace("20px")`

   - the weekly or monthly number of cases of many diseases exhibit often a seasonal pattern as well as short term dependence

`r vspace("10px")`

   - the underlying daily level of atmospheric pollutants, e.g. PM $_{10}$, will show strong correlation over consecutive days
    because their lifetime lasts over several days

`r vspace("20px")`

- To the contrary of spatial models, there is a natural order to any time series data which is used in specifying the models.

---

# Spatial patterns

1\. Data
  - Disease counts $y_{i}$ in area $i$ and stratum $k$, aggregated over a time period, $i=1,\ldots,N$, $k=1,\ldots,K$
  `r vspace("10px")`
  - Population counts $n_{ik}$ in area $i$ and stratum $k$, aggregated over a time period
  `r vspace("10px")`
  - Expected numbers $E_i = \sum_k n_{ik} r_k$, where $r_k$ reference rate for stratum (age, gender,...)


2\. Spatial smoothing using BYM model

\begin{align*}
y_i & \sim  \hbox{Poisson}(E_i \rho_i); \;\;\; i=1,...,N\\
\log \rho_i & =  b_0 + v_i + u_i\\
v_i &\sim  \hbox{Normal}(0, \sigma^2_v)\\
{\bm u} & \sim   \hbox{ICAR}({\bm W}, \sigma_u^2) \rightarrow  u_i | u_{j, \;\; j\ne i}  \sim  \hbox{Normal}\left(\frac{\sum_{j} w_{ij} u_j}{\sum_{j} w_{ij}}, \sigma^2_u/n_i\right)
\end{align*}

with $w_{ij}=1$ if areas $i$ and $j$ are neighbours, 0 otherwise

---

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  ???
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  b_0 + \beta t \quad \hbox{ simple linear regression}\\
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  b_0 + \beta t \quad \hbox{ simple linear regression}\\
& =   b_0 + \psi_t \quad \hbox{ global temporal smoothing}\\
& \psi_t \sim \hbox{Normal}(0, \sigma^2_{\psi})\\
\end{align*}


---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  b_0 + \beta t \quad \hbox{ simple linear regression}\\
& =   b_0 + \psi_t \quad \hbox{ global temporal smoothing}\\
& \psi_t \sim \hbox{Normal}(0, \sigma^2_{\psi})\\
& =   b_0 + \gamma_t \quad \hbox{ local temporal smoothing}\\
&    \gamma_t \sim \hbox{distribution ?}\\
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  b_0 + \beta t \quad \hbox{ simple linear regression}\\
& =   b_0 + \psi_t \quad \hbox{ global temporal smoothing}\\
& \psi_t \sim \hbox{Normal}(0, \sigma^2_{\psi})\\
& =   b_0 + \gamma_t \quad \hbox{ local temporal smoothing}\\
&    \gamma_t \sim \hbox{distribution ?}\\
& =   b_0 + \gamma_t + \psi_t \quad \hbox{ global and local temporal smoothing}\\
\end{align*}


---

# Ohio Lung cancer data

- Data on lung cancer in 88 counties of Ohio, 1968-1988
- Annual rates adjusted by gender and race

```{r ohio_data, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.5, fig.width = 3, fig.align = "center"}
library(sf)
library(dplyr)
library(ggplot2)
library(SpatialEpi)
ohio.gen <- read_sf("~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/Ohio Lung Cancer/tl_2010_39_county00.shp")
ohio.data <- read.table("~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/Ohio Lung Cancer/Data.txt", header=T)
ohio.data <- as_tibble(ohio.data)
#Expected calculation
E=list()
for(i in 1:21){
E[[i]] <- expected(ohio.data$n[ohio.data$year==i],ohio.data$y[ohio.data$year==i],4)
}
E <- as_tibble(unlist(E))
ohio.data <- ohio.data %>% group_by(county, year) %>%  summarise(y=sum(y),n=sum(n)) 
ohio.data = data.frame(ohio.data, E=E)

ohio.data <- ohio.data %>%  mutate(E=value, SMR=y/E) %>% select(-value)

ohio.data %>% group_by(year) %>%  summarise(sum.y=sum(y), sum.E=sum(E)) %>% mutate(mean.SMR=sum.y/sum.E) %>%  ggplot(aes(year,mean.SMR)) + 
  geom_line() + labs(y = "Lung Cancer SMR", x = "Year")
```
---

# Ohio Lung cancer data: modelled temporal trend

- Smoothed temporal trend using random walk models (posterior mean) seen in the previous lecture

```{r ohio-rw, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.5, fig.width = 3, fig.align = "center"}
formula.rw1 <- sum.y~1 + f(year, model="rw1")
formula.rw2 <- sum.y~1 + f(year, model="rw2")
data.fit <- ohio.data %>% group_by(year) %>%  summarise(sum.y=sum(y), sum.E=sum(E)) %>% mutate(mean.SMR=sum.y/sum.E)

library(INLA)
mod.rw1 <- inla(formula.rw1, data=data.fit, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))
mod.rw2 <- inla(formula.rw2, data=data.fit, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))

data.fit %>% mutate(rw1=mod.rw1$summary.fitted.values[,1], rw2=mod.rw2$summary.fitted.values[,1]) %>% ggplot(aes(year, log(rw1)))  + geom_line(aes(year, log(rw2)), linetype = "dashed") + geom_point(aes(year, log(mean.SMR)), size=0.3) + theme(legend.position = "none") + labs(y = "Lung Cancer Rate", x = "Year")
```

---

# Ohio Lung cancer spatial pattern (no temporal dimension)

```{r ohio-space, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.8, fig.width = 6, fig.align = "center"}
formula.bym <- sum.y~1 + f(county, model="bym2", graph="~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/Ohio Lung Cancer/Ohio.graph")
data.fit2 <- ohio.data %>% group_by(county) %>%  summarise(sum.y=sum(y),sum.E=sum(E)) %>% mutate(mean.SMR= sum.y/sum.E)
mod.bym <- inla(formula.bym, data=data.fit2, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))

data.out <- data.fit2 %>% mutate(bym=mod.bym$summary.fitted.values[,1], COUNTYFP00=ohio.gen$COUNTYFP00)

out_map <- left_join(ohio.gen,data.out, by="COUNTYFP00")

map1 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = mean.SMR)) + 
            scale_fill_viridis_c(direction=-1) + theme(legend.position = "bottom") +
            guides(fill=guide_legend(title="SMR")) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))


map2 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = bym)) + 
            scale_fill_viridis_c(direction=-1) + theme(legend.position = "bottom") +
            guides(fill=guide_legend(title=expression(rho[i]))) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))

library(gridExtra)
grid.arrange(map1,map2,ncol=2)
```
---

name: space-to-time
  
`r vspace("250px")`

.myblue[.center[.huge[
**From space to space-time**]]]

---



# Disease mapping: Extending space to space-time

- Disease mapping is usually carried out on aggregated data over a time period

`r vspace("10px")`

- Rather than suppressing the time dimension, it can be interesting to use models that combine the space and time dimension

`r vspace("10px")`

- The stability (or not) of the spatial pattern can aid interpretation

`r vspace("10px")`

- The specific space-time components of the model can potentially pinpoint
unusual/emerging hazards

`r vspace("10px")`

- Data:

  - $y_{it}$ and
  - $\hbox{E}_{it}$: the observed and expected number of cases in area $i$ at time $t$ calculated as $E_{it} = \sum_k n_{itk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

---

# Schematic representation I

`r include_fig("representation_ST1.jpg")`

---

# Linear spatio-temporal model

- A simple model assumes a linear effect of time:

$$\log\rho_{it} = b_0 + u_i + v_i + \beta \times t$$

  - Main spatial effect $u_i + v_i$
  - Main linear trend $\beta$ (global time effect) 


- A differential effect $\delta_i$ can be added

$$\log\rho_{it} = b_0 + u_i + v_i + (\beta + \delta_i) \times t$$

  - If $\delta_i < 0$ then the area-specific trend is less steep than the mean trend, whilst $\delta_i > 0$ implies that the area-specific trend is steeper than the mean trend. 


---

# Dynamic spatio-temporal model

`r include_fig("representation_ST.jpg", width="80%")`

---

# Ohio Lung cancer - temporal SMRs

Log SMR time trends in each county

  - Slightly increasing trend but lots of variation across the counties
  
```{r ohio-ST, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.5, fig.width = 6, fig.align = "center"}
ggplot()  +  geom_line(data=ohio.data,aes(year, log(SMR), group=county)) + geom_line(data=data.fit,aes(year, log(mean.SMR), colour="red")) + theme(legend.position = "none") + labs(y = "Lung Cancer Rate", x = "Year")
```

---

# Simple linear spatio-temporal model (Model 1)

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + \beta*t + u_i + v_i\\
\end{align*}
where

- $b_0$ overall log RR in Ohio over the 21-year period

- $v_i \sim \hbox{Normal}(0, \sigma^2_v)$ spatially unstructured RE

- $\mathbf{u} \sim \hbox{ICAR}(\mathbf{W}, \sigma^2_u)$ spatially structured RE

- $\exp(\beta)$ is the change in the RR associated with a 1-year increase in time

```{r formula_mod1, echo=TRUE, eval=TRUE}
formula.mod1 = y ~ 1 + f(county,model="bym",
                   graph=Ohio.adj) + year
```

---

# Log-linear temporal model with unstructured temporal RE (Model 2)

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + \beta*t + u_i + v_i + \color{red}{\psi_t}\\
\end{align*}

where

- $b_0$ overall log RR in Ohio over the 21-year period

- $v_i \sim \hbox{Normal}(0, \sigma^2_v)$ spatially unstructured RE

- $\mathbf{u} \sim \hbox{ICAR}(\mathbf{W}, \sigma^2_u)$ spatially structured RE

- $\exp(\beta)$ is the change in the RR associated with a 1-year increase in time

- $\color{red}{\psi_t  \sim \hbox{Normal}(0, \sigma^2_{\psi})}$ .red[temporally unstructured RE]

```{r formula_mod2, echo=TRUE, eval=TRUE}
year2 = ohio.data$year
formula.mod2 = y ~ 1 + f(county,model="bym",
                  graph=Ohio.adj) + 
                  f(year2,model="iid") + 
                  year
```
---

# Ohio lung cancer - comparison models 1 and 2

```{r echo=FALSE, eval=TRUE}
Ohio.adj="~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/Ohio Lung Cancer/Ohio.graph"
```

```{r echo=TRUE, eval=TRUE}
mod1 = inla(data=ohio.data,formula=formula.mod1, E=E, family="poisson")
mod2 = inla(data=ohio.data,formula=formula.mod2, E=E, family="poisson")
# Posterior mean and  95% CI for Model 1
mod1$summary.fixed

#Posterior mean and 95% CI for Model 2
mod2$summary.fixed
```

---

# Ohio lung cancer - comparison models 1 and 2

Let's look at the temporal random effect in model 2
```{r echo=FALSE, eval=TRUE, fig.height= 2, fig.width=5}
ggplot() + geom_line(aes(x=seq(1968,1988),y=mod2$summary.random$year2$mean)) + 
  theme(axis.title.x = element_blank()) + ylab(expression(log(psi[t]))) 

```

Some temporal structure which suggests that the linear term is not capturing the temporal trend well

---

# Simple additive space-time structure version 1 (Model 3)

This model assumes that the space time variations can be captured
by the superimposition of a BYM spatial model and a structured time
trend

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it})\\
\log\rho_{it} & =  b_0 + b_i + \class{red}{\gamma_t}\\
\end{align*}

- $b_i = v_i + u_i$ 
- $v_i \sim \hbox{Normal}(0, \sigma^2_v)$ spatially unstructured RE
- $\mathbf{u} \sim \hbox{ICAR}(\mathbf{W}, \sigma^2_u)$ spatially structured RE
- $\class{red}{\gamma_t  \sim \hbox{RW}(1)}$ temporally structured RE with variance parameter $\sigma^2_{\gamma}$
- Assuming a BYM2 specification, there are 3 hyperparameters

  - $\tau_b$, the marginal precision of $b_i$
  - $\phi$, the proportion of spatial variability explained by the spatially structured component 
  - $\sigma^2_\gamma$, the conditional variance of the RW(1) modelling the time trend

```{r formula_mod3, echo=TRUE, eval=TRUE}
formula.mod3 = y ~ 1 + f(county,model="bym",
                  graph=Ohio.adj) + 
                  f(year,model="rw1") 
```

```{r mod3, echo=TRUE, eval=TRUE}
mod3 = inla(data=ohio.data,formula=formula.mod3, E=E, family="poisson")
```
---

# Simple additive space-time structure version 2 (Model 4)

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + u_i + v_i + \class{red}{\psi_t} + \class{red}{\gamma_t}\\
\end{align*}

where
- $b_0$ overall log RR in Ohio over the 21-year period
- $v_i \sim \hbox{Normal}(0, \sigma^2_v)$ spatially unstructured RE
- $\mathbf{u} \sim \hbox{ICAR}(\mathbf{W}, \sigma^2_u)$ spatially structured RE
- $\class{red}{\gamma_t  \sim \hbox{RW}(1)}$ temporally structured RE with variance parameter $\sigma^2_{\gamma}$
- $\class{red}{\psi_t  \sim \hbox{Normal}(0, \sigma^2_{\psi})}$ temporally unstructured RE

```{r formula_mod4, echo=TRUE, eval=TRUE}
formula.mod4 = y ~ 1 + f(county,model="bym",
                   graph=Ohio.adj) + 
                   f(year,model="rw1") +   
                   f(year2,model="iid")
```
```{r mod4, echo=TRUE, eval=TRUE}
mod4 = inla(data=ohio.data,formula=formula.mod4, E=E, family="poisson")
```

---

# Ohio lung cancer - comparison models 3 and 4

Posterior means of the temporal RE

Model 3 = $\gamma_t$ 

Model 4 = $\psi_t + \gamma_t$

`r include_fig("ohio11.jpeg",width="40%")`


---

# Ohio lung cancer - area-specific temporal RR

Posterior means of the temporal RE $\gamma_t$ (model 3)

`r include_fig("ohio10.jpeg", width="50%")`

---

# Ohio lung cancer - spatial patterns (all models)

Posterior means of the spatial random effects for the 4 different models

```{r echo=FALSE, eval=TRUE, fig.width=6, fig.height=2.5, fig.align="center"}
ohio.out <- data.frame(county=seq(1,88),mod1=mod1$summary.random$county$mean[1:88],mod2=mod1$summary.random$county$mean[1:88],
mod3=mod3$summary.random$county$mean[1:88],mod4=mod4$summary.random$county$mean[1:88])
ohio.out <- as_tibble(ohio.out)
breaks =c(-4,-1,-0.1,0.1,1,4)
ohio.out <- ohio.out %>% mutate(mod1_cat = cut(mod1,breaks, include.lowest=TRUE),
                                mod2_cat = cut(mod2,breaks, include.lowest=TRUE),
                                mod3_cat = cut(mod3,breaks, include.lowest=TRUE),
                                mod4_cat = cut(mod4,breaks, include.lowest=TRUE)
)

ohio.gen$county <- seq(1,88)
out_map <- left_join(ohio.gen,ohio.out, by="county")


map1 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = mod1_cat)) + 
            theme_bw() + scale_fill_brewer(palette = "OrRd") +             guides(fill=guide_legend(title="model 1")) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))

map2 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = mod2_cat)) + 
            theme_bw() + scale_fill_brewer(palette = "OrRd") +             guides(fill=guide_legend(title="model 2")) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))

map3 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = mod3_cat)) + 
            theme_bw() + scale_fill_brewer(palette = "OrRd") +             guides(fill=guide_legend(title="model 3")) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))

map4 <- ggplot() + geom_sf(data = out_map, 
                   aes(fill = mod4_cat)) + 
            theme_bw() + scale_fill_brewer(palette = "OrRd") + 
            guides(fill=guide_legend(title="model 4")) +
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))

library(gridExtra)
grid.arrange(map1,map2,map3, map4, ncol=2)
```

Similar patterns with the 4 models
---

# Space-time model with exchangeable interactions (Model 5)

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + b_i + \;\gamma_t + \psi_t + \color{blue}{\delta_{it}}\\
b_i & =  v_i + u_i\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  \hbox{N}(0,\sigma^2_{\psi})\\
\color{blue}{\delta_{it}} & \color{blue}{\sim  \hbox{Normal}(0, \sigma^2_{\delta})}
\end{align*}

The simple additive model 4 can be extended by including **space-time interactions parameters**, $\delta_{it}$, modelled as
exchangeable random effects, that capture departure from the
additive structure
---

# Interpretation of the interactions in model 5
- The interactions $\delta_{it}$ allow to highlight unusual temporal trends

- Rules based on the posterior probabilities $p(\delta_{it} > 0)$ for at least 1 time $t$

- Ohio lung cancer:  6 counties with unusual temporal trends

.pull-left[
"usual" temporal trends

`r include_fig("ohio14.jpeg", width="70%")`
]

.pull-right[
unusual temporal trends
`r include_fig("ohio13.jpeg", width="70%")`
]


---

# INLA code for model with interaction (Model 5)

- In INLA is very easy to include the interaction in the `formula` environment

- We need to specify an index for the interaction, i.e. for each combination of area/time  

\begin{align*}
\log\rho_{it} & =  b_0 + b_i + \gamma_t + \psi_t + \delta_{it}\\
b_i & =  v_i + u_i\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  N(0,\sigma^2_{\psi})\\
\delta_{it}  \sim & \hbox{Normal}(0, \sigma^2_{\delta})
\end{align*}

```{r st_mod, echo=TRUE, eval=FALSE}
formula.intI = y ~ + f(county,model="bym",
                       graph=Ohio.adj) +
                      f(year,model="rw1") +
                      f(year2,model="iid") +
                      f(area.year,model="iid")
```


---

name: interactions
  
`r vspace("250px")`

.myblue[.center[.huge[
**Types of interactions**]]]

---





# Another example: Birth weight in Georgia

- Count of babies weigthing less than 2500g in 159 counties in Georgia (US)
- Period 2000 – 2010

```{r Georgia_SMR, echo=FALSE, eval=TRUE, out.width="55%", opts=list(width="55%")}
library(tidyverse)
Georgia_map <- read_sf("~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/LBW in Georgia/co13_d00.shp")
Georgia_map1<-subset(Georgia_map,AREA>0.02)

#Georgia_map[Georgia_map$CO13_D00_I %in% c(100,105,98,137),]$NAME==c("Macon1","Taylor1","Taylor2","Lee1")
Georgia_data <- as_tibble(read.csv("~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/LBW in Georgia/data.final.csv"))
#add_duplicates <- Georgia_data[Georgia_data$X %in% c("Taylor", "Lee" ,   "Macon"),]
#Georgia_data <- Georgia_data %>%  add_row(add_duplicates)

Georgia_data <- Georgia_data %>% mutate(NAME=X,ID.area=seq(1,159), SMR2000=obs2000/Exp2000,SMR2001=obs2001/Exp2001,SMR2002=obs2002/Exp2002,
                        SMR2003=obs2003/Exp2003,SMR2004=obs2004/Exp2004,SMR2005=obs2005/Exp2005,
                        SMR2006=obs2006/Exp2006,SMR2007=obs2007/Exp2007,SMR2008=obs2008/Exp2008,
                        SMR2009=obs2009/Exp2009,SMR2010=obs2010/Exp2010)
out_map <- inner_join(Georgia_map1,Georgia_data,by="NAME")
out_map <- out_map %>%  mutate(ID=seq(1,159))
yearly_SMR <- out_map %>%
  pivot_longer(cols = starts_with("SMR"),
               values_to = "SMR",
               names_to = "year_SMR") %>%
select(NAME, ID, SMR, year_SMR, geometry) %>% mutate(ID.area = ID, ID.year=as.numeric(as.factor(year_SMR)),ID.area.year=seq(1,1749))

library(scales)
ggplot() +
  geom_sf(data = yearly_SMR , aes(fill = SMR), color = "black", size = 0.25) + 
  scale_fill_distiller(name="SMR", palette = "BrBG", breaks = pretty_breaks(5))+
  facet_wrap(~ year_SMR, ncol = 4) +
  labs(title="")
```

---

# Different types of interactions `r Cite(bibfile, "knorr2000bayesian")`

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log \rho_{it} & =  b_0 + b_i  + \gamma_t + \psi_t + \color{blue}{\delta_{it}}\\
b_i & = v_i + u_i\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  N(0,\sigma^2_{\psi})\\
\end{align*}

```{r, results = "asis", echo = FALSE, message = FALSE, eval=TRUE}
Interaction <- c("I", "II", "III", "IV")
Parameters <- c(paste("\\(v\\)", "and", "\\(\\psi\\)"),
         paste("\\(v\\)", "and", "\\(\\gamma\\)"),
         paste("\\(u\\)", "and", "\\(\\psi\\)"),
         paste("\\(u\\)", "and", "\\(\\gamma\\)"))
Rank <- c("nT","n(T-1) for RW1, n(T-2) for RW2", "(n-1)T","(n-1)(T-1) for RW1, (n-1)(T-2) for RW2")
DF<- as_tibble(cbind(Interaction, Parameters, Rank))
kable(DF, col.names = c("Interaction", "Parameters","Rank"),escape=F, caption = "Characteristics of ST interaction") %>%
  kable_styling(latex_options = "hold_position")
```

- We will learn how to develop Type I interaction in INLA
---

# How to model interactions

- Data in R format
```{r dataGeorgia, echo=FALSE, eval=TRUE} 
data_obs = Georgia_data %>%
  pivot_longer(cols = starts_with("obs"),
               values_to = c("Obs"),
               names_to = c("year_obs")) %>%   
select(NAME, ID.area, year_obs, Obs) %>% mutate(ID.year=as.numeric(as.factor(year_obs)))

data_exp = Georgia_data %>%
  pivot_longer(cols = starts_with("Exp"),
               values_to = c("Exp"),
               names_to = c("year_exp")) %>%   
select(NAME, ID.area, year_exp, Exp) %>% mutate(ID.year=as.numeric(as.factor(year_exp)))

data_INLA = left_join(data_obs,data_exp,by=c("ID.area","ID.year"))

data_INLA = data_INLA %>% mutate(y=Obs,E=Exp,NAME=NAME.x,ID.area.year=seq(1,dim(data_INLA)[1])) %>%  select(!c(NAME.x,NAME.y, year_obs, year_exp))

data_INLA
```

`r vspace("-10px")`

- We need to make sure to have an index for area (`ID.area`), one for time (`ID.year`) and one for the interaction (`ID.area.year`)

---

# Type I interaction in INLA

Let's adopt Type I interaction, which assumes that the two unstructured effects $v_i$ and $\psi_t$ interact

`r vspace("-10px")`


```{r echo=FALSE, eval=TRUE}
Georgia.adj = "~/Dropbox/TEACHING/YEAR_2023/Advanced_Analytics_2023/Material/Session4.2/Datasets/LBW in Georgia/Georgia.graph"
ID.year2 = data_INLA$ID.year
formula.intI = y ~ + f(ID.area,model="bym2",
                       graph=Georgia.adj) +
                     f(ID.year,model="rw1") +
                    f(ID.year2,model="iid") +                                    f(ID.area.year,model="iid")
```

```{r INLA_ST1, echo=TRUE, eval=FALSE}
Georgia.adj = "Georgia.graph"
ID.year2 = data_INLA$ID.year
formula.intI = y ~ + f(ID.area,model="bym2",
                       graph=Georgia.adj) +
                     f(ID.year,model="rw1") +
                     f(ID.year2,model="iid") +
                     f(ID.area.year,model="iid")
```

`r include_fig("INLA_ST1_mod-1.png", width="40%")`

```{r INLA_ST1_mod, echo=FALSE, eval=FALSE}
mod.intI = inla(formula.intI, data=data_INLA, E=E, family="poisson", control.predictor = list(compute = TRUE), control.compute=list(dic=TRUE, waic=TRUE))
res_intI<-tibble(ID.area.year=mod.intI$summary.random$ID.area.year$ID,interaction=mod.intI$summary.random$ID.area.year$mean)
res_intI <- left_join(yearly_SMR,res_intI, by="ID.area.year")
res_intI$int_cat <- cut(res_intI$interaction,  breaks=c(-0.1,-0.05, -0.01,
                  0.01, 0.05, 0.1),include.lowest = T)
ggplot() +
geom_sf(data = res_intI, aes(fill = int_cat))  + theme_bw() +  scale_fill_brewer(palette = "PuOr") +
            guides(fill=guide_legend(title=NULL)) +
            theme(text = element_text(size=20),
                  axis.text.x = element_blank(),
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.year, ncol = 4, labeller=labeller(ID.year=c("1"="2000","2"="2001","3"="2002","4"="2003","5"="2004","6"="2005","7"="2006","8"="2007","9"="2008","10"="2009","11"="2010"))) +
  labs("")
```
---

# Kronecker product

- For the interactions of type II-IV we will need to use the **Kronecker product** to specify the dependencies. It is a matrix multiplication which returns a block matrix.

For instance
$$\left[\begin{array}{cc}
1 & 2 \\ 3 & 4
\end{array}\right] \otimes
\left[ \begin{array}{cc}
 0 & 5 \\ 6 & 7
 \end{array} \right] = \left[
\begin{array}{cc}
1 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] & 2 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] \\
3 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] & 4 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right]
\end{array}\right]$$

- There is a function in `R` which does exactly that: `kronecker(a,b)`

- We will not go through type II-IV of interactions, as they are pretty complex and also they can take a **very long** time to run

- If you are interested in knowing more how to set interactions using the Kronecker product in `INLA` see `r Citet(bibfile,"goicoa2018spatio")`
---


# Summary

`r vspace("20px")`

- Increase quality of datasets that are both spatially and temporally indexed

`r vspace("20px")`

- Advanced methods to deal with this type of data

`r vspace("20px")`

- Allow to interpret the stability (or not) of the spatial patterns

---

# References

```{r refs, echo=FALSE, results="asis"}
PrintBibliography(bibfile,.opts=list(max.names=3))
```

