---
title: "Session 8.2: Spatio-temporal modelling for area data"
params: 
   conference: "Imperial College London"
   date: ""
   short_title: "MSc in Epidemiology"
output:
   xaringan::moon_reader: 
    includes: 
       # This line adds a logo based on the format selected in the file 'assets/include_logo.html'
       in_header: "assets/latex_macros.html" 
       # NB: the actual options (eg placement of the logo and actual logo file) can be changed there
     #  after_body: "assets/insert-logo.html" # Monica commented this
    seal: false
    yolo: no
    lib_dir: libs
    nature:
      beforeInit: ["assets/remark-zoom.js","https://platform.twitter.com/widgets.js"]
      navigation:
        scroll: false # disable slide transitions by scrolling
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
      titleSlideClass:
      - center
      - middle
    self_contained: false 
    css:
    - "assets/beamer.css"
editor_options: 
  chunk_output_type: console
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 999)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
#options(htmltools.preserve.raw = FALSE)
# https://stackoverflow.com/questions/65766516/xaringan-presentation-not-displaying-html-widgets-even-when-knitting-provided-t
```

```{r echo=F, message=FALSE, warning=FALSE, comment=NA}
source("assets/setup.R")
library(INLA)
xaringanExtra::use_panelset()
#bibfile=RefManageR::ReadBib("~/Dropbox/TEACHING/YEAR_2024/Bayes_Spatial_2024/Material/Session8.2/Biblio.bib",check = FALSE)
bibfile=RefManageR::ReadBib("C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Session8.2/Biblio.bib",check = FALSE)
```

class: title-slide

# `r rmarkdown::metadata$title``r vspace("10px")` `r rmarkdown::metadata$subtitle`

## `r rmarkdown::metadata$author`

### `r rmarkdown::metadata$institute`    

### `r rmarkdown::metadata$params$conference`

<!-- Can also separate the various components of the extra argument 'params', eg as in 
### `r paste(rmarkdown::metadata$params, collapse=", ")`
-->

`r ifelse(is.null(rmarkdown::metadata$params$date),format(Sys.Date(),"%e %B %Y"),rmarkdown::metadata$params$date)`

---

layout: true  

.my-footer[ 
.alignleft[ 
&nbsp; &copy; Blangiardo | Pirani | Gascoigne
]
.aligncenter[
`r rmarkdown::metadata$params$short_title` 
]
.alignright[
`r rmarkdown::metadata$params$conference`, `r short_date` 
]
] 

```{css,echo=FALSE, eval=FALSE}
.red {
  color: red;
}
.blue {
  color: 0.14 0.34 0.55;
}

.content-box-blue { background-color: #F0F8FF; }

}
```

---

# Learning Objectives

At the end of this session you should be able to:

`r vspace("10px")`


- Explain how to extend spatial or temporal models to spatio-temporal models 


`r vspace("10px")`


- Fit Bayesian space-time models using `R-INLA`


`r vspace("20px")`


The topics covered in this lecture can be found in Chapter 7 of the book Spatial and **Spatio-Temporal Bayesian Models with `R-INLA`**.  

---

# Outline 

`r vspace("30px")`

1\. [Temporal dependence](#temporal-structure)

`r vspace("30px")`

2\. [From space to space-time](#space-to-time)

`r vspace("30px")`

3\. [Type of interactions](#interactions)

---

name: temporal-structure
  
`r vspace("250px")`

.myblue[.center[.huge[
**Temporal dependence**]]]

---

# Temporal dependence

- Similarly to spatial dependence, it is sometimes necessary to model temporal dependence of data or of parameters:

`r vspace("20px")`

   - the weekly or monthly number of cases of many diseases exhibit often a seasonal pattern as well as short term dependence

`r vspace("10px")`

   - the underlying daily level of atmospheric pollutants, e.g. PM $_{10}$, will show strong correlation over consecutive days
    because their lifetime lasts over several days

`r vspace("20px")`

- To the contrary of spatial models, there is a natural order to any time series data which is used in specifying the models.

---

# Spatial patterns

1\. Data
  - Disease counts $y_{i}$ in area $i$ and stratum $k$, aggregated over a time period, $i=1,\ldots,N$, $k=1,\ldots,K$
  `r vspace("10px")`
  - Population counts $n_{ik}$ in area $i$ and stratum $k$, aggregated over a time period
  `r vspace("10px")`
  - Expected numbers $E_i = \sum_k n_{ik} r_k$, where $r_k$ reference rate for stratum (age, gender,...)


2\. Spatial smoothing using BYM model

\begin{align*}
y_i & \sim  \hbox{Poisson}(E_i \rho_i); \;\;\; i=1,...,N\\
\eta_i = \log \rho_i & =  b_0 + b_i\\
\boldsymbol{b} &= \frac{1}{\sqrt{\tau_b}}(\sqrt{1-\phi}\boldsymbol{v}_{*} + \sqrt{\phi}\boldsymbol{u}_{*})
\end{align*}

---

# Ohio Lung cancer spatial pattern 


- Data on lung cancer in 88 counties of Ohio, 1968-1988

- Purely spatial analysis

```{r ohio-space, echo=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(SpatialEpi)
ohio.gen <- read_sf("C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/Ohio Lung Cancer/tl_2010_39_county00.shp")
ohio.data <- read.table("C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/Ohio Lung Cancer/Data.txt", header=T)
#ohio.gen <- read_sf("~/Dropbox/TEACHING/YEAR_2024/Bayes_Spatial_2024/Material/Datasets/Ohio Lung Cancer/tl_2010_39_county00.shp")
#ohio.data <- read.table("~/Dropbox/TEACHING/YEAR_2024/Bayes_Spatial_2024/Material/Datasets/Ohio Lung Cancer/Data.txt", header=T)

ohio.data <- as_tibble(ohio.data)

#Expected calculation
E=list()
for(i in 1:21){
E[[i]] <- expected(ohio.data$n[ohio.data$year==i],ohio.data$y[ohio.data$year==i],4)
}
E <- as_tibble(unlist(E))
ohio.data <- ohio.data %>% group_by(county, year) %>%  summarise(y=sum(y),n=sum(n)) 
ohio.data = data.frame(ohio.data, E=E)

ohio.data <- ohio.data %>%  mutate(E=value, SMR=y/E) %>% select(-value)
Ohio.adj="C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/Ohio Lung Cancer/Ohio.graph"
#Ohio.adj="~/Dropbox/TEACHING/YEAR_2024/Bayes_Spatial_2024/Material/Datasets/Ohio Lung Cancer/Ohio.graph"

formula.bym <- sum.y~1 + f(county, model="bym2", graph=Ohio.adj)
data.fit2 <- ohio.data %>% group_by(county) %>%  summarise(sum.y=sum(y),sum.E=sum(E)) %>% mutate(mean.SMR= sum.y/sum.E)
mod.bym <- inla(formula.bym, data=data.fit2, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))

data.out <- data.fit2 %>% mutate(bym=mod.bym$summary.fitted.values[,1], COUNTYFP00=ohio.gen$COUNTYFP00)

out_map <- left_join(ohio.gen,data.out, by="COUNTYFP00")
```

```{r eval=TRUE, include=TRUE, fig.height = 2.6, fig.width = 6, fig.align = "center"}
ggplot() + geom_sf(data = out_map, 
                   aes(fill = bym)) + scale_colour_continuous(type = "viridis") + 
          theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(),  axis.ticks.y=element_blank(),
          legend.key.size = unit(0.4, 'cm'), legend.text = element_text(size=6),
          legend.title = element_text(size=6))
```
---

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

Note that $E_t$ makes sense if the time series covers a long period, e.g. years, otherwise it is common to assign $E_t = \sum_t y_t / T$

`r vspace("20px")`

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  ???
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

Note that $E_t$ makes sense if the time series covers a long period, e.g. years, otherwise it is common to assign $E_t = \sum_t y_t / T$

`r vspace("20px")`

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =  b_0 + \beta t \quad \hbox{ simple linear regression}\\
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

Note that $E_t$ makes sense if the time series covers a long period, e.g. years, otherwise it is common to assign $E_t = \sum_t y_t / T$

`r vspace("10px")`

OR

`r vspace("10px")`

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =   b_0 + \psi_t \quad \hbox{ global temporal smoothing}\\
& \psi_t \sim \hbox{Normal}(0, \sigma^2_{\psi})\\
\end{align*}


---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

Note that $E_t$ makes sense if the time series covers a long period, e.g. years, otherwise it is common to assign $E_t = \sum_t y_t / T$

`r vspace("10px")`

OR

`r vspace("10px")`

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =   b_0 + \gamma_t \quad \hbox{ local temporal smoothing}\\
&    \gamma_t \sim RW(\sigma^2_{\gamma})\\
\end{align*}

---

count:false

# Temporal trends

1\. Data

  - Disease counts $y_{tk}$ in time period $t$ and stratum $k$, aggregated over space,  $t=1,\ldots,T$ (equally-spaced time intervals), $k=1,\ldots,K$ `r vspace("10px")`
  - Population counts $n_{tk}$ in time period $t$ and stratum $k$, aggregated over space 
`r vspace("10px")`
  - Expected numbers $E_t = \sum_k n_{tk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

Note that $E_t$ makes sense if the time series covers a long period, e.g. years, otherwise it is common to assign $E_t = \sum_t y_t / T$

`r vspace("10px")`

OR

`r vspace("10px")`

2\. Temporal trends

\begin{align*}
y_t & \sim  \hbox{Poisson}(E_t \rho_t); \;\;\; t=1,...,T\\
\log \rho_t & =   b_0 + \gamma_t + \psi_t \quad \hbox{ global and local temporal smoothing}\\
& \psi_t \sim \hbox{Normal}(0, \sigma^2_{\psi})\\
&    \gamma_t \sim RW(\sigma^2_{\gamma})\\
\end{align*}


---

# Ohio Lung cancer data: modelled temporal trend

- Annual rates adjusted by gender and race

- Fitting different temporal trends (linear, .red[local smoothing], .blue[local+global smoothing])

```{r ohio-rw, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.5, fig.width = 3, fig.align = "center"}
data.fit <- ohio.data %>% group_by(year) %>%  summarise(sum.y=sum(y), sum.E=sum(E)) %>% mutate(mean.SMR=sum.y/sum.E)
data.fit$year1<-data.fit$year

formula.linear <- sum.y~1 + year
formula.rw1 <- sum.y~1 + f(year, model="rw1")
formula.convolution <- sum.y~1 + f(year, model="rw1") + f(year1,model="iid")

library(INLA)
mod.linear <- inla(formula.linear, data=data.fit, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))
mod.rw1 <- inla(formula.rw1, data=data.fit, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))
mod.convolution <- inla(formula.convolution, data=data.fit, E=sum.E, family="poisson", control.predictor = list(compute = TRUE))

data.fit %>% mutate(linear=mod.linear$summary.fitted.values[,1], rw1=mod.rw1$summary.fitted.values[,1], conv=mod.convolution$summary.fitted.values[,1]) %>% ggplot(aes(year, linear))  + geom_line(aes(year, linear), linetype = "solid", colour = "black") + geom_line(aes(year, rw1), linetype = "dashed", colour = "red") + geom_line(aes(year, conv), linetype = "dotted", colour = "blue") + geom_point(aes(year, mean.SMR), size=0.3) + theme(legend.position = "none") + labs(y = "Lung Cancer posterior RR", x = "Year")
```

---



name: space-to-time
  
`r vspace("250px")`

.myblue[.center[.huge[
**From space to space-time**]]]

---



# Disease mapping: Extending space to space-time

- Disease mapping is usually carried out on aggregated data over a time period

`r vspace("10px")`

- Rather than suppressing the time dimension, it can be interesting to use models that combine the space and time dimension

`r vspace("10px")`

- The stability (or not) of the spatial pattern can aid interpretation

`r vspace("10px")`

- The specific space-time components of the model can potentially pinpoint
unusual/emerging hazards

`r vspace("10px")`

- Data:

  - $y_{it}$ 
  `r vspace("10px")`
  
  - $\hbox{E}_{it}$: the observed and expected number of cases in area $i$ at time $t$ calculated as $E_{it} = \sum_k n_{itk} r_k$, where $r_k$ reference rate for stratum (age, gender,...)

---

# Dynamic spatio-temporal model

`r include_fig("representation_ST.jpg", width="80%")`

---

# Ohio Lung cancer - temporal SMRs

Log SMR time trends in each county

  - Slightly increasing trend but lots of variation across the counties
  
```{r ohio-ST, echo=FALSE, eval=TRUE, include=TRUE, fig.height = 2.5, fig.width = 6, fig.align = "center"}
ggplot()  +  geom_line(data=ohio.data,aes(year, log(SMR), group=county)) + geom_line(data=data.fit,aes(year, log(mean.SMR), colour="red")) + theme(legend.position = "none") + labs(y = "Lung Cancer SMR", x = "Year")
```

---

# Model 1: linear time + space

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + \beta \times t + b_i\\
\end{align*}
where

- $b_0$ overall log RR in Ohio over the 21-year period

- $b_i$ is the weighted average of spatially structured (ICAR) and unstructured random effects  

- $\exp(\beta)$ is the change in the RR associated with a 1-year increase in time

```{r formula_mod1, echo=TRUE, eval=TRUE}
formula.mod1 = y ~ 1 + year + f(county,model="bym2",
                   graph=Ohio.adj) 
```

```{r mod1, echo=TRUE, eval=TRUE}
mod1 = inla(data=ohio.data,formula=formula.mod1, E=E, family="poisson", control.compute=list(waic=TRUE))
```

Additional temporal random effects can be added...

---

# Model 2: linear time + space + structured time

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + \beta\times t + b_i + \color{red}{\gamma_t}\\
\end{align*}

where

- $b_0$ overall log RR in Ohio over the 21-year period

- $b_i$ is the weighted average of spatially structured (ICAR) and unstructured random effects  

- $\exp(\beta)$ is the change in the RR associated with a 1-year increase in time

- Looking back at the purely temporal analysis it seems a random walk was able to explain the local changes in time $\color{red}{\gamma_t  \sim \hbox{Normal}(\gamma_{t-1}, \sigma^2_{\gamma})}$ 

```{r formula_mod2, echo=TRUE, eval=TRUE}
year2<-ohio.data$year
formula.mod2 = y ~ 1  + 
                  year + f(county,model="bym2",
                  graph=Ohio.adj) + 
                  f(year2,model="rw1")
```

```{r mod2, echo=TRUE, eval=TRUE}
mod2 = inla(data=ohio.data,formula=formula.mod2, E=E, family="poisson", control.compute=list(waic=TRUE))
```
---

# Comparing model 1 and 2: Fixed effects

```{r comp, echo=TRUE, eval=TRUE}
mod1$summary.fixed
mod2$summary.fixed
```

---

# Comparing model 1 and 2: Fitted values

```{r comp2, echo=FALSE, include=TRUE, eval=TRUE,  fig.height = 2, fig.width = 5, fig.align = "center"}
data.out <- ohio.data %>% mutate(linear=mod1$summary.fitted.values[,1], rw1=mod2$summary.fitted.values[,1])
 
data.out  %>% group_by(year) %>% summarise(mean.lin=mean(linear), mean.rw=mean(rw1)) %>%  ggplot(aes(x=year))  + geom_line(aes(year, mean.lin), linetype = "solid", colour = "black")  + geom_line(aes(year, mean.rw), linetype = "solid", colour = "red") + 
scale_colour_manual("", breaks = c("mean.lin", "mean.rw"), values = c("black", "red")) + labs(y="Posterior means")
```

---
# Model 3: space + structured time + unstructured time

- Exploring if a combination of random effects could be more flexible in capturing the temporal trend than a linear trend

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + b_i + \color{red}{\gamma_t} + \color{red}{\psi_t}\\
\end{align*}

where

- $b_0$ overall log RR in Ohio over the 21-year period

- $b_i$ is the weighted average of spatially structured (ICAR) and unstructured random effects  

- The temporal pattern is now modelled with two random effects (global + local smoothing) 

```{r formula_mod3, echo=TRUE, eval=TRUE}
formula.mod3 = y ~ 1  + 
                  f(county,model="bym2", graph=Ohio.adj) + 
                  f(year,model="rw1") +
                  f(year2, model="iid")
```

```{r mod3, echo=TRUE, eval=TRUE}
mod3 = inla(data=ohio.data,formula=formula.mod3, E=E, family="poisson", control.compute=list(waic=TRUE))
```
---

# Comparing model 1 and 3: Fixed effects

```{r comp4, echo=TRUE, eval=TRUE}
mod1$summary.fixed
mod3$summary.fixed
```

---

# Comparing model 2 and 3: Fitted values

```{r comp5, echo=FALSE, include=TRUE, eval=TRUE, fig.height = 2.5, fig.width = 6, fig.align = "center"}
data.out <- ohio.data %>% mutate(linear=mod1$summary.fitted.values[,1], conv=mod3$summary.fitted.values[,1])
 
data=data.out  %>% group_by(year)  %>%  summarise(linear=mean(linear),rw1=mean(conv)) %>%  mutate(mod2=mod2$summary.fixed[1,1]+mod2$summary.fixed[2,1]*seq(1,21) + mod2$summary.random$year2[,2], mod3=mod3$summary.fixed[1,1] + mod3$summary.random$year[,2]+mod3$summary.random$year2[,2])

ggplot(data,aes(x=year))  + geom_line(aes(year, mod2), linetype = "solid", colour = "black") + geom_line(aes(year, mod3), linetype = "dashed", colour = "red") + labs(y="Posterior means")
```
---

# Do we need space-time interactions in spatio-temporal models?

- When we have temporal random effects, if we look at area specific trends this is what we would see:

`r include_fig("ohio10.jpeg", width="40%")`

.red[Parallel lines as the temporal trend is assumed to be the same across all the spatial units]

---


# Space-time model with exchangeable interactions

- It is easy to allow for spatio-temporal interactions, which add flexibility to the model

- A general specification of the model will be

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log\rho_{it} & =  b_0 + b_i + \;\gamma_t + \psi_t + \color{blue}{\delta_{it}}\\
b_i & =  v_i + u_i\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  \hbox{N}(0,\sigma^2_{\psi})\\
\color{blue}{\delta_{it}} & \color{blue}{\sim  \hbox{Normal}(0, \sigma^2_{\delta})}
\end{align*}

The **space-time interaction parameter**, $\delta_{it}$, is modelled as
exchangeable random effects, that capture departure from the
additive structure (given by the BYM2 in space and RW1+iid in time).

---

# Interpretation of the interactions 

- The interactions $\delta_{it}$ allow to highlight unusual temporal trends

- Rules based on the posterior probabilities $p(\delta_{it} > 0)$ for at least 1 time $t$

.pull-left[
"usual" temporal trends

`r include_fig("ohio14.jpeg", width="70%")`
]

.pull-right[
unusual temporal trends
`r include_fig("ohio13.jpeg", width="70%")`
]


---

# INLA code for model with interaction

- In INLA is very easy to include the interaction in the `formula` environment

- We need to specify an index for the interaction, i.e. for each combination of area/time  

\begin{align*}
\log\rho_{it} & =  b_0 + b_i + \gamma_t + \psi_t + \delta_{it}\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  N(0,\sigma^2_{\psi})\\
\delta_{it}  \sim & \hbox{Normal}(0, \sigma^2_{\delta})
\end{align*}

```{r st_mod, echo=TRUE, eval=FALSE}
formula.intI = y ~ + f(county,model="bym2",
                       graph=Ohio.adj) +
                      f(year,model="rw1") +
                      f(year2,model="iid") +
                      f(area.year,model="iid")
```


---

name: interactions
  
`r vspace("250px")`

.myblue[.center[.huge[
**Types of interactions**]]]

---

# Another example: Birth weight in Georgia

- Count of babies weigthing less than 2500g in 159 counties in Georgia (US)
- Period 2000 – 2010

```{r Georgia_SMR, echo=FALSE, eval=TRUE, out.width="55%", opts=list(width="55%")}
library(tidyverse)
#Georgia_map <- read_sf("~/Dropbox/TEACHING/YEAR_2024/Bayes_Spatial_2024/Material/Datasets/LBW in Georgia/co13_d00.shp")
Georgia_map <- read_sf("C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/LBW in Georgia/co13_d00.shp")
Georgia_map1<-subset(Georgia_map,AREA>0.02)

#Georgia_map[Georgia_map$CO13_D00_I %in% c(100,105,98,137),]$NAME==c("Macon1","Taylor1","Taylor2","Lee1")
Georgia_data <- as_tibble(read.csv("C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/LBW in Georgia/data.final.csv"))
#add_duplicates <- Georgia_data[Georgia_data$X %in% c("Taylor", "Lee" ,   "Macon"),]
#Georgia_data <- Georgia_data %>%  add_row(add_duplicates)

Georgia_data <- Georgia_data %>% mutate(NAME=X,ID.area=seq(1,159), SMR2000=obs2000/Exp2000,SMR2001=obs2001/Exp2001,SMR2002=obs2002/Exp2002,
                        SMR2003=obs2003/Exp2003,SMR2004=obs2004/Exp2004,SMR2005=obs2005/Exp2005,
                        SMR2006=obs2006/Exp2006,SMR2007=obs2007/Exp2007,SMR2008=obs2008/Exp2008,
                        SMR2009=obs2009/Exp2009,SMR2010=obs2010/Exp2010)
out_map <- inner_join(Georgia_map1,Georgia_data,by="NAME")
out_map <- out_map %>%  mutate(ID=seq(1,159))
yearly_SMR <- out_map %>%
  pivot_longer(cols = starts_with("SMR"),
               values_to = "SMR",
               names_to = "year_SMR") %>%
select(NAME, ID, SMR, year_SMR, geometry) %>% mutate(ID.area = ID, ID.year=as.numeric(as.factor(year_SMR)),ID.area.year=seq(1,1749))

library(scales)
ggplot() +
  geom_sf(data = yearly_SMR , aes(fill = SMR), color = "black", size = 0.25) + 
  scale_fill_distiller(name="SMR", palette = "BrBG", breaks = pretty_breaks(5))+
  facet_wrap(~ year_SMR, ncol = 4) +
  labs(title="")
```

---

# Different types of interactions `r Cite(bibfile, "knorr2000bayesian")`

\begin{align*}
y_{it} & \sim  \hbox{Poisson}(\hbox{E}_{it} \rho_{it}) \\
\log \rho_{it} & =  b_0 + b_i  + \gamma_t + \psi_t + \color{blue}{\delta_{it}}\\
b_i & = v_i + u_i\\
\gamma_t & \sim  \hbox{RW(1)}\\
\psi_t & \sim  N(0,\sigma^2_{\psi})\\
\end{align*}

```{r, results = "asis", echo = FALSE, message = FALSE, eval=TRUE}
Interaction <- c("I", "II", "III", "IV")
Parameters <- c(paste("Unstructured in space ", "and", "Unstructured in time"),
         paste("Unstructured in space ", "and", "RW in time"),
         paste("ICAR in space ", "and", "Unstructured in time"),
         paste("ICAR in space ", "and", "RW in time"))
Rank <- c("nT","n(T-1) for RW1, n(T-2) for RW2", "(n-1)T","(n-1)(T-1) for RW1, (n-1)(T-2) for RW2")
DF<- as_tibble(cbind(Interaction, Parameters, Rank))
kable(DF, col.names = c("Interaction", "Parameters","Rank"),escape=F, caption = "Characteristics of ST interaction") %>%
  kable_styling(latex_options = "hold_position")
```

---

# How to model interactions

- Data in R format
```{r dataGeorgia, echo=FALSE, eval=TRUE} 
data_obs = Georgia_data %>%
  pivot_longer(cols = starts_with("obs"),
               values_to = c("Obs"),
               names_to = c("year_obs")) %>%   
select(NAME, ID.area, year_obs, Obs) %>% mutate(ID.year=as.numeric(as.factor(year_obs)))

data_exp = Georgia_data %>%
  pivot_longer(cols = starts_with("Exp"),
               values_to = c("Exp"),
               names_to = c("year_exp")) %>%   
select(NAME, ID.area, year_exp, Exp) %>% mutate(ID.year=as.numeric(as.factor(year_exp)))

data_INLA = left_join(data_obs,data_exp,by=c("ID.area","ID.year"))

data_INLA = data_INLA %>% mutate(y=Obs,E=Exp,NAME=NAME.x,ID.area.year=seq(1,dim(data_INLA)[1])) %>%  select(!c(NAME.x,NAME.y, year_obs, year_exp))

data_INLA
```

`r vspace("-10px")`

- We need to make sure to have an index for area (`ID.area`), one for time (`ID.year`) and one for the interaction (`ID.area.year`)

---

# Type I interaction in INLA

Let's adopt Type I interaction, which assumes that the two unstructured effects $v_i$ and $\psi_t$ interact

`r vspace("-10px")`


```{r echo=FALSE, eval=TRUE}
Georgia.adj = "C:/Users/magb/Dropbox/Bayes_Spatial_2025/Material/Datasets/LBW in Georgia/Georgia.graph"
ID.year2 = data_INLA$ID.year
formula.intI = y ~ + f(ID.area,model="bym2",
                       graph=Georgia.adj) +
                     f(ID.year,model="rw1") +
                    f(ID.year2,model="iid") +                                    f(ID.area.year,model="iid")
```

```{r INLA_ST1, echo=TRUE, eval=FALSE}
Georgia.adj = "Georgia.graph"
ID.year2 = data_INLA$ID.year
formula.intI = y ~ + f(ID.area,model="bym2",
                       graph=Georgia.adj) +
                     f(ID.year,model="rw1") +
                     f(ID.year2,model="iid") +
                     f(ID.area.year,model="iid")
```

`r include_fig("INLA_ST1_mod-1.png", width="40%")`

```{r INLA_ST1_mod, echo=FALSE, eval=FALSE}
mod.intI = inla(formula.intI, data=data_INLA, E=E, family="poisson", control.predictor = list(compute = TRUE), control.compute=list(dic=TRUE, waic=TRUE))
res_intI<-tibble(ID.area.year=mod.intI$summary.random$ID.area.year$ID,interaction=mod.intI$summary.random$ID.area.year$mean)
res_intI <- left_join(yearly_SMR,res_intI, by="ID.area.year")
res_intI$int_cat <- cut(res_intI$interaction,  breaks=c(-0.1,-0.05, -0.01,
                  0.01, 0.05, 0.1),include.lowest = T)
ggplot() +
geom_sf(data = res_intI, aes(fill = int_cat))  + theme_bw() +  scale_fill_brewer(palette = "PuOr") +
            guides(fill=guide_legend(title=NULL)) +
            theme(text = element_text(size=20),
                  axis.text.x = element_blank(),
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.year, ncol = 4, labeller=labeller(ID.year=c("1"="2000","2"="2001","3"="2002","4"="2003","5"="2004","6"="2005","7"="2006","8"="2007","9"="2008","10"="2009","11"="2010"))) +
  labs("")
```
---

# Kronecker product

- For the interactions of type II-IV we will need to use the **Kronecker product** to specify the dependencies. It is a matrix multiplication which returns a block matrix.

For instance
$$\left[\begin{array}{cc}
1 & 2 \\ 3 & 4
\end{array}\right] \otimes
\left[ \begin{array}{cc}
 0 & 5 \\ 6 & 7
 \end{array} \right] = \left[
\begin{array}{cc}
1 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] & 2 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] \\
3 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right] & 4 \left[\begin{array}{cc}
0 & 5\\
6 & 7
\end{array}\right]
\end{array}\right]$$

- There is a function in `R` which does exactly that: `kronecker(a,b)`

- Type II-IV of interactions are pretty complex and can take a **very long** time to run

- We breifly discuss, but for more information on how to set interactions using the Kronecker product in `INLA` see `r Citet(bibfile,"goicoa2018spatio")`
---

# Type II interaction: in INLA

- Type II combines the structured temporal main effect $\gamma_t$ and the unstructured
spatial effect $v_i$.

- We use first create the structure matrix for the time component assuming a random walk of order 1
```{r eval=TRUE, echo=TRUE}
#RW1
D1 <- diff(diag(11),differences=1)
Q.gammaRW1 <- t(D1)%*%D1
#Let's look at it
Q.gammaRW1[1:5,1:5]
```

---

# Type II interaction: in INLA

- Then we create the Kronecker product  $v \otimes$ $\gamma$ and note from the table in slide 35 that the rank deficiency is n (159 in our case)
```{r eval=TRUE, echo=TRUE}
#Kronecker product (for RW1)
R <- kronecker(diag(159),Q.gammaRW1)
R[1:5,1:5]
```

---

# Type II interaction: in INLA

- Finally we need to impose some more constraints as the matrix have spatio-temporal interaction matrix has a rank deficiency

- We calculate the eigenvalues of the structure matrix of the interaction term and the extra constraints are those where the eigenvalues are zero

```{r eval=TRUE, echo=TRUE}
eigenQ <- eigen(R)
ids <- which(eigenQ$values < 0.00001)
cMat <- t(eigenQ$vectors[,ids])
```

The `formula` environment needs to be changed to
```{r eval=TRUE, echo=TRUE}
formula.intII<- y ~ f(ID.area,model="bym2",graph=Georgia.adj) +
                    f(ID.year,model="rw1") +
                    f(ID.year2,model="iid") +
                    f(ID.area.year, model="generic0", Cmatrix=R, rankdef = nrow(cMat),
                    extraconstr = list(A = cMat, e = rep(0, nrow(cMat))))
```

---

# Type II interaction: in INLA

`r include_fig("INLA_ST2_mod-1.png", width="60%")`

```{r INLA_ST2_mod, echo=FALSE, eval=FALSE}
mod.intII <- inla(formula.intII, data=data_INLA, E=E, family="poisson", control.predictor = list(compute = TRUE), control.compute=list(dic=TRUE, waic=TRUE))
res_intII<-tibble(ID=data_INLA$ID.area,ID.year= data_INLA$ID.year,ID.area.year=mod.intII$summary.random$ID.area.year$ID,interaction=mod.intII$summary.random$ID.area.year$mean)

#res_intII<-tibble(ID=ID.area.int,ID.year= ID.year.int,ID.area.year=mod.intII$summary.random$ID.area.int$ID,interaction=mod.intII$summary.random$ID.area.int$mean)
res_intII <- left_join(out_map,res_intII, by="ID")
res_intII$int_cat <- cut(res_intII$interaction,  breaks=c(-0.2,-0.05, -0.01,
                  0.01, 0.05, 0.2),include.lowest = T)

ggplot() +
  geom_sf(data = res_intII, aes(fill = int_cat)) + theme_bw() +  scale_fill_brewer(palette = "PuOr") +
            guides(fill=guide_legend(title=NULL)) +
            theme(text = element_text(size=20),
                  axis.text.x = element_blank(),
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.year, ncol = 4, labeller=labeller(ID.year=c("1"="2000","2"="2001","3"="2002","4"="2003","5"="2004","6"="2005","7"="2006","8"="2007","9"="2008","10"="2009","11"="2010"))) +
  labs("")
```

---

# Type III interaction in INLA

- Type III combines the spatially structured main effect $u_i$ and the unstructured temporal effect $\psi_t$.

- We use first create the structure matrix for the spatial component using the neighborhood graph
```{r eval=TRUE, echo=TRUE}
g <- inla.read.graph(Georgia.adj)
Q.xi <- matrix(0, g$n, g$n)
for (i in 1:g$n){
  Q.xi[i,i]=g$nnbs[[i]]
  Q.xi[i,g$nbs[[i]]]=-1
}
Q.xi[1:5,1:5]
```

---

# Type III interaction in INLA

- Then we create the Kronecker product $u$ $\otimes$ $\psi$ and note from the table in slide 35 that the rank deficiency is T (11 in our case)
```{r eval=TRUE, echo=TRUE}
#Kronecker product (for RW1)
R <- kronecker(Q.xi,diag(11))
R[1:5,1:15]
```
---

# Type III interaction in INLA

- Finally we need to impose some more constraints (on the eigenvalues equal to 0)

```{r eval=TRUE, echo=TRUE}
eigenQ <- eigen(R)
ids <- which(eigenQ$values < 0.00001)
cMat <- t(eigenQ$vectors[,ids])
```

The `formula` environment becomes
```{r eval=TRUE, echo=TRUE}
formula.intIII<- y ~ f(ID.area,model="bym2",graph=Georgia.adj) +
                    f(ID.year,model="rw1") +
                    f(ID.year2,model="iid") +
                    f(ID.area.year, model="generic0", Cmatrix=R, rankdef = nrow(cMat),
                     extraconstr = list(A = cMat, e = rep(0, nrow(cMat))))
```
---

# Type III interaction: in INLA

`r include_fig("INLA_ST3_mod-1.png", width="60%")`

```{r INLA_ST3_mod, echo=FALSE, eval=FALSE}
mod.intIII <- inla(formula.intIII, data=data_INLA, E=E, family="poisson", control.predictor = list(compute = TRUE), control.compute = list(dic=TRUE, waic=TRUE))
res_intIII<-tibble(ID=data_INLA$ID.area,ID.year= data_INLA$ID.year,ID.area.year=mod.intIII$summary.random$ID.area.year$ID,interaction=mod.intIII$summary.random$ID.area.year$mean)

#res_intII<-tibble(ID=ID.area.int,ID.year= ID.year.int,ID.area.year=mod.intII$summary.random$ID.area.int$ID,interaction=mod.intII$summary.random$ID.area.int$mean)
res_intIII <- left_join(out_map,res_intIII, by="ID")
res_intIII$int_cat <- cut(res_intIII$interaction,  breaks=c(-0.02,-0.005, -0.001,
                  0.001, 0.005, 0.02),include.lowest = T)
ggplot() +
  geom_sf(data = res_intIII, aes(fill = int_cat)) + theme_bw() +  scale_fill_brewer(palette = "PuOr") +
            guides(fill=guide_legend(title=NULL)) +
            theme(text = element_text(size=20),
                  axis.text.x = element_blank(),
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.year, ncol = 4, labeller=labeller(ID.year=c("1"="2000","2"="2001","3"="2002","4"="2003","5"="2004","6"="2005","7"="2006","8"="2007","9"="2008","10"="2009","11"="2010"))) +
  labs("")
```

---

# Type IV interaction in INLA

- Type IV is the most complex type of interaction, assuming that the spatially and temporally structured effects $u_i$ and $\gamma_t$ interact
-  We assume that the temporal dependency structure for each area is not independent from all the other areas anymore, but depends on the temporal pattern of the neighboring areas as well


- We create the Kronecker product $u$ $\otimes$ $\gamma$ and note from the table in slide 35 that the rank deficiency is n+T-1 (169 in our case)
```{r eval=TRUE, echo=TRUE}
#Kronecker product (for RW1)
R <- kronecker(Q.xi,Q.gammaRW1)
eigenQ <- eigen(R)
ids <- which(eigenQ$values < 0.00001)
cMat <- t(eigenQ$vectors[,ids])
```

The `formula` environment becomes
```{r eval=TRUE, echo=TRUE}
formula.intIV<- y ~ f(ID.area,model="bym2",graph=Georgia.adj) +
                    f(ID.year,model="rw1") +
                    f(ID.year,model="iid") +
                    f(ID.area.year, model="generic0", Cmatrix=R, rankdef = nrow(cMat),
                    extraconstr = list(A = cMat, e = rep(0, nrow(cMat))))
```
---

# Type IV interaction: in INLA

`r include_fig("INLA_ST4_mod-1.png", width="60%")`

```{r INLA_ST4_mod, echo=FALSE, eval=FALSE}
mod.intIV <- inla(formula.intIV, data=data_INLA, E=E, family="poisson", control.predictor = list(compute = TRUE), control.compute = list(dic=TRUE, waic=TRUE))
res_intIV<-tibble(ID=data_INLA$ID.area,ID.year= data_INLA$ID.year,ID.area.year=mod.intIV$summary.random$ID.area.year$ID,interaction=mod.intIV$summary.random$ID.area.year$mean)

#res_intII<-tibble(ID=ID.area.int,ID.year= ID.year.int,ID.area.year=mod.intII$summary.random$ID.area.int$ID,interaction=mod.intII$summary.random$ID.area.int$mean)
res_intIV <- left_join(out_map,res_intIV, by="ID")
res_intIV$int_cat <- cut(res_intIV$interaction,  breaks=c(-0.2,-0.05, -0.01,
                  0.01, 0.05, 0.2),include.lowest = T)
ggplot() +
  geom_sf(data = res_intIV, aes(fill = int_cat)) + theme_bw() +  scale_fill_brewer(palette = "PuOr") +
            guides(fill=guide_legend(title=NULL)) +
            theme(text = element_text(size=20),
                  axis.text.x = element_blank(),
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.year, ncol = 4, labeller=labeller(ID.year=c("1"="2000","2"="2001","3"="2002","4"="2003","5"="2004","6"="2005","7"="2006","8"="2007","9"="2008","10"="2009","11"="2010"))) +
  labs("")
```

---

# Model selection

It might be useful to employ indexes like DIC and WAIC to decide which model is the most appropriate for the problem at hand

In this example:

```{r eval=TRUE, echo=FALSE}
WAIC <-c(11602,11569,11659,11612)
DF$WAIC <- WAIC
kable(DF, col.names = c("Interaction", "Parameters","Rank", "WAIC"),escape=F, caption = "Model fitting for the different ST models") %>%
  kable_styling(latex_options = "hold_position")
```

So we conclude that the model with the interaction of order II is the most appropriate for this data.

---


# Summary

`r vspace("20px")`

- Increase quality of datasets that are both spatially and temporally indexed

`r vspace("20px")`

- Advanced methods to deal with this type of data

`r vspace("20px")`

- Allow to interpret the stability (or not) of the spatial patterns

`r vspace("20px")`

--

- Model selection tools are useful to choose which interaction to use 

`r vspace("20px")`

- Careful with interactions of order above I as they can take a **very long** time to run

---

# References

```{r refs, echo=FALSE, results="asis"}
PrintBibliography(bibfile,.opts=list(max.names=3))
```

