---
title: "Practical 1 - Bayes rule"
author: "Spatial and Spatio-Temporal Bayesian Models with `R-INLA`"
output: 
  html_document:
    css: practical.css
    toc: true
    toc_float: true
---   

\pagenumbering{gobble} 
\pagenumbering{arabic} 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.align = "center", class.source='klippy')
```
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'),color = 'darkred',
               tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

# 1. Comparing between the prior (unconditional) and the posterior (conditional)

For each scenario below, you are given a pair of events, $A$ and $B$. Explain what you believe to be the relationships between the prior and posterior probabilities of $B$. For example, $P(B|A) < P(B)$.

  a. $A$ is *you live in a house with someone who has COVID-19* and $B$ is *you become ill with COVID-19*
  b. $A$ is *you wear a mask in areas with poor ventilation* and $B$ is *you become ill with COVID-19*
  c. $A$ is *you attend a cervical cancer screening and have the HPV vaccine* and $B$ is *you contract cervical cancer*
  d. $A$ is *you regularly smoke cigarettes* and $B$ is *you contract cervical cancer*
  e. $A$ is *socially distance when you have flu-like symptoms* and $B$ is *you contract cervical cancer*
  
# 2. Probabilities

Consider the following events: $A$ is *going to an underground bar*, $B$ is *testing positive for COVID-19* and, $C$ is *having COVID-19*. Write down the following statements in probabilistic notation.

  a. $47\%$ of people that go to an underground bar get COVID-19
  b. If you have COVID-19, you have an $86\%$ chance of a positive COVID-19 test
  c. $39\%$ of the population have COVID-19
  d. If you do not have COVID-19, you have an $4\%$ chance of a negative COVID-19 test
  e. All your friends have said they would go to the underground bar if they have a negative COVID-19 test
  f. $1\%$ of people go to an underground bar and have a positive COVID-19 test

  
# 3. Cervical cancer example
  
The prevalence of cervical cancer in a population is $8\%$. A pap smear test is used to identify those who have cervical cancer and is known to be $57\%$ sensitive and $76\%$ specific. In the following, let $A$ be the event *an individual has cervical cancer* and $B$ be the event *an individual tests positive to having cervical cancer on the pap smear test*. 

Using the information above, answer the following:
  
  a. What are the probabilities that an individual has and has not got cervical cancer?
  b. What are the probabilities that an individual returns a positive pap smear test **given** they have cervical cancer and that an individual returns a negative pap smear test **given** they do not have cervical cancer?
  c. Consider the following table
  
```{r, echo = FALSE}
library(tidyverse)
data.frame(Event = c('$A$', '$A^c$', '$Total$'), B1 = c('', '', ''), B2 = c('', '' , ''), Total = c('', '', 1)) %>%
  kableExtra::kable(., align = 'cccc', col.names = c('$Event$', '$B$', '$B^c$', '$Total$'))  %>%
  kableExtra::column_spec(1:4, border_left = T, border_right = T) %>% 
  kableExtra::kable_styling()
```

  using your answers from the part a and b , calculate the following:
  
  i. The probability an individual has cervical cancer **and** returns a positive pap smear test
  ii. The probability an individual has cervical cancer **and** returns a negative pap smear test
  iii. The probability an individual does not have cervical cancer **and** returns a positive pap smear test
  iv. The probability an individual does not have cervical cancer **and** returns a negative pap smear test
  v. The probability an individual returns a negative pap smear test
  vi. The probability an individual returns a positive pap smear test
  vii. Complete the table
  viii. Finally return the probability that an individual has cervical cancer given that they test positive on the pap smear.
  
# 4. New screening method

An alternative screening method for cervical cancer is to use a colposcopy which is know to be $92\%$ and $67\%$ sensitive and specific. 

  a. Using the same prevalence as in Question 2,  obtain the probability of having cervical cancer if you test positive at the colposcopy. Compare the pros and cons of the two different screening methods.
  b. What does it mean for a test to be highly sensitive?
  c. What does it mean for a test to be highly specific?
  d. Given the two screening methods in Questions 3 and 4, what is the best method of screening for cervical cancer? You can use, none, either, or both.




  