---
title: "Session 2.1: Bayesian inference"
params: 
   conference: "Bayesian modelling for Spatial and Spatio-temporal data"
   location: "Imperial College"
   date: ""
   short_title: "MSc in Epidemiology"

output:
  xaringan::moon_reader: 
    includes: 
       in_header: "../assets/latex_macros.html" 
       # This line adds a logo based on the format selected in the file 'assets/include_logo.html'
       # NB: the actual options (eg placement of the logo and actual logo file) can be changed there
       #after_body: "../assets/insert-logo.html"
    seal: false
    yolo: no
    lib_dir: libs
    nature:
      beforeInit: ["https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
      titleSlideClass:
      - center
      - middle
    self_contained: false 
    css:
    - "../assets/beamer.css"
---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 999)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
```
```{r echo=F,message=FALSE,warning=FALSE,comment=NA}
# Sources the R file with all the relevant setup and commands
source("assets/setup.R")
library(INLA)

# Stuff from 'xaringanExtra' (https://pkg.garrickadenbuie.com/xaringanExtra)
# This allows the use of panels (from 'xaringanExtra')
xaringanExtra::use_panelset()
# This allows to copy code from the slides directly
#xaringanExtra::use_clipboard()
# This freezes the frame for when there's a gif included
#xaringanExtra::use_freezeframe()

# Defines the path to the file with the .bib entries (in case there are references)
bibfile=ReadBib("~/Dropbox/Bayes_Spatial_2023/Material/Biblio.bib",check = FALSE)
#bibfile=ReadBib("~/Dropbox/Lavori condivisi/2015_Book/ShortCourse/VIBASS/Biblio.bib",check = FALSE)

```

class: title-slide

# `r rmarkdown::metadata$title``r vspace("10px")` `r rmarkdown::metadata$subtitle`

## `r rmarkdown::metadata$author`

### `r rmarkdown::metadata$institute`    

### `r rmarkdown::metadata$params$conference`, `r rmarkdown::metadata$params$location` 

<!-- Can also separate the various components of the extra argument 'params', eg as in 
### `r paste(rmarkdown::metadata$params, collapse=", ")`
-->

`r ifelse(is.null(rmarkdown::metadata$params$date),format(Sys.Date(),"%e %B %Y"),rmarkdown::metadata$params$date)`

---

layout: true  

.my-footer[ 
.alignleft[ 
&nbsp; &copy; Marta Blangiardo | Monica Pirani 
]
.aligncenter[
`r rmarkdown::metadata$params$short_title` 
]
.alignright[
`r rmarkdown::metadata$params$conference`, `r short_date` 
]
] 

```{css,echo=FALSE, eval=FALSE}
.red {
  color: red;
}
.blue {
  color: 0.14 0.34 0.55;
}

.content-box-blue { background-color: #F0F8FF; }

}
```
<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>


---

# Learning Objectives

After this session you should be able to:

`r vspace("10px")`

- Understand how Bayes Theorem can be applied to random variables

`r vspace("10px")`

- Describe what conjugacy means

`r vspace("10px")`

- Obtain posterior distribution for the Beta-Binomial and Gamma-Poisson families

`r vspace("10px")`

The topics treated in this lecture are presented in Chapter 3  of the book `r Citet(bibfile, "blangiardo2015spatial")` and in Chapter 2.3, 3.1-3.4 and 5.2-5.3 of the book `r Citet(bibfile, "johnson2022bayes")`.

---

# Outline 

1\. [Bayes Theorem for random variables](#BayRandom)

`r vspace("30px")`


2\. [A quick recap](#Recap)

`r vspace("30px")`

3\. [What is conjugacy?](#Conjugacy)

`r vspace("30px")`

4\. [Some conjugacy models: Beta-Binomial](#Beta-Binomial)

`r vspace("30px")`

5\. [Some conjugacy models: Gamma-Poisson](#PoissonGamma)


---

name: BayRandom

`r vspace("250px")`

.myblue[.center[.huge[
**Bayes Theorem for random variables**]]]


---

# An example

- A clinical trial is carried out to assess the efficacy of a preventive treatment for migraine
`r vspace("20px")`
- A group of 20 patients are offered the new drug and they have to report if they have a migraine episode in the next week after taking the medication.
`r vspace("20px")`
- Our aim is to estimate the probability of success of this new drug $(\theta)$
`r vspace("20px")`
- Our data here is $Y$: the number of patients reporting no migraine episodes.
`r vspace("20px")`
Note that $Y$ is a **random variable** (look at recording 3 for a more detailed description of this concept)

---

# Prior probability model

As a first step we need to assign a prior on $\theta$. 

As a simplification let's assume that $\theta$ is discrete and can only get thevalues 0.1,0.4,0.8 with the following probability function, which specifies the prior probability of each possible $\theta$ value:
`r vspace("30px")`

```{r tableAlt}
library(kableExtra)
df<- data.frame(theta=c(0.1,0.4,0.8), p=c(0.2,0.5,0.3))
df %>%  kbl(col.names = c("\\(\\theta\\)", "\\(p(\\theta)\\)")) %>%  kable_classic(full_width=F,  position = "center")
```

`r vspace("30px")`

Note that this prior reflects some sort of information from a previous study on a similar compound and put 50% probability on the event that 40% of the patients will report a reduction in migraine.


---

# The Binomial data model

- Our random variable $Y$ can take any discrete value between 0 and 20 (total number of patients)

- It will depend on the probability $\theta$

- For our formal Bayesian analysis, we must model this dependence of $Y$ on $\theta$ through a **conditional probability model**  

- We make two assumptions about the trials: (1) the outcome of any one patient doesn’t influence the outcome of another; and (2) the probability of success does not change among patients

- We can use the Binomial model


.content-box-green[

$Y \sim \mbox{Binomial}(n, \theta)$

with conditional probability function
$$p(y \mid \theta) = \frac{n!}{y!(n-y)!}\theta^{y}(1-\theta)^{n-y}$$
- Mean of this distribution is $E(Y) = n \theta$
- Variance is $V(Y) = n \theta(1-\theta)$

.red[ Check out recording 4 for a recap on the Binomial distribution]
]

---

# The Binomial data model

This model allows to calculate ANY conditional probability. For instance, conditioning on $\theta=0.4$ let's see the difference in the probability of getting 12 or 15 successes

1. $$p(y=12 \mid \theta=0.4) = \frac{20!}{12!8!}0.4^{12}(1-0.4)^{8}= 0.035$$ 
2. $$p(y=15 \mid \theta=0.4) = \frac{20!}{15!5!}0.4^{15}(1-0.4)^{5} = 0.0013$$

-- 

Note you can get these results in R using

```{r eval=TRUE, echo=TRUE}
#1
dbinom(12,20,0.4)
#2
dbinom(15,20,0.4)
```
---

# Binomial likelihood function

- The Binomial provides a theoretical model of the data we might observe. 
- In the end we observe 10 successes out of the 20 patients $(y=10)$.
- The next step in our Bayesian analysis is to determine how compatible this particular data is with the various possible $\theta$

.content-box-green[
We need to evaluate the *likelihood* of getting 10 successes in the trial under each possible value of $\theta$
]

Similarly to last week with events, the likelihood function follows from evaluating the conditional probability function $p(y=10\mid \theta)$ for all the possible $\theta$ values:

1. $$p(y=10 \mid \theta=0.1) = \frac{20!}{10!10!}0.1^{10}(1-0.1)^{10}= 0.00000644$$ 
2. $$p(y=10 \mid \theta=0.4) = \frac{20!}{10!10!}0.4^{10}(1-0.4)^{10} = 0.117$$
3. $$p(y=10 \mid \theta=0.8) = \frac{20!}{10!10!}0.8^{10}(1-0.8)^{10} = 0.002$$

---

# Likelihood vs probability function

Let's recall the fundamental difference between probability function and likelihood function

`r vspace("50px")`

.pull-left[
.content-box-yellow[
When $\theta$ is known, the **conditional probability function** $p(\cdot\mid \theta)$ allows us to compare the probabilities of different values of the data $Y$ occurring with $\theta$:
$$p(y_1\mid \theta) \text{  vs  } p(y_2\mid \theta)$$
]
]

.pull-right[
.content-box-yellow[
When $Y$ is known, the **likelihood function** $L(\cdot\mid y) = p(y \mid \cdot)$ allows us to compare the relative likelihood of the data $y$ under different possible values of $\theta$:
$$L(\theta_1 \mid y) = p(y \mid \theta_1) \text{  vs  } L(\theta_2 \mid y) = p(y \mid \theta_2)$$
]
]

`r vspace("50px")`

So $L(\cdot \mid y)$ provides the tool we need to evaluate the relative compatibility of data $Y=y$ with various $\theta$ values.

---

# Normalising constant

- Now we have a prior for $\theta$ and a likelihood and as Bayesian we want to **balance** these two pieces of information to obtain the posterior.

- Something is missing...

--

The normalising constant!

This is the total probability of having $y=10$ successes. How do we get this?

--

.content-box-green[
**Law of total probability**: 
$$P(A) = \sum_j P(A \mid B_j)P(B_j) \text{   (for events)}$$
$$P(Y=y) = \sum_j p(Y \mid \theta_j)p(\theta_j) \text{   (for discrete probability functions)}$$
]

`r vspace("20px")`


So we get $$P(Y=10) = 0.00000644 \times 0.2 +  0.117 \times 0.5 + 0.002 \times 0.3 = 0.059$$

Interpretation: across all the possible $\theta$ there is only around 6% chance that there are 10 successes.
---

# Posterior distribition

Finally we are ready to apply Bayes theorem and get the posterior distribution

$$p(\theta \mid y=10) = \frac{p(\theta)L(\theta \mid y=10)}{p(y=10)} \text{  for  } \theta \in \{0.1,0.4,0.8\}$$
which gives us:

$$p(\theta=0.1 \mid y=10) = \frac{0.2 \times 0.0000064}{0.059} = 0$$
$$p(\theta=0.4 \mid y=10) = \frac{0.5 \times 0.117}{0.059} = 0.99$$

$$p(\theta=0.8 \mid y=10) = \frac{0.3 \times 0.002}{0.059} = 0.01$$
---

# Comparing prior and posterior
.pull-left[
```{r table1, echo=FALSE, eval=TRUE}
library(kableExtra)
df<- data.frame(theta=c(0.1,0.4,0.8), p=c(0.2,0.5,0.3), post=c(0,0.99,0.01), lik=c(0.0000064,0.117,0.002))
df[,1:3] %>%  kbl(col.names = c("\\(\\theta\\)", "\\(p(\\theta)\\)", "\\(p(\\theta \\mid y=10)\\)")) %>%  kable_classic(full_width=F,  position = "center")
```
]

.pull-right[

```{r eval=TRUE, fig.height = 13, fig.width = 14, fig.show='hold',opts=list(width="110%")}
library(gridExtra)
library(ggplot2)
p1<-ggplot(df, aes(df[,1],df[,2])) + geom_point()  + geom_segment(aes(x=df[,1],y=0,xend=df[,1],yend=df[,2])) + xlab(expression(theta)) + ylab(expression(p(theta))) + ggtitle("Prior") + ylim(0,1) + theme(axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold", hjust=0.5))

p2<-ggplot(df, aes(df[,1],df[,3])) + geom_point()  + geom_segment(aes(x=df[,1],y=0,xend=df[,1],yend=df[,3])) + xlab(expression(theta)) + ylab(expression(paste("p(",theta, "| y=10)"))) + ggtitle("Posterior") + 
  theme(axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold", hjust=0.5))
grid.arrange(p1,p2,nrow=1)
```
]
---

# Posterior shortcut

.pull-left[
- Moving forward we can actually forget about calculating the normalising constant

- Note that on slide 12 the $p(y=10)$ appears at the denominator of all the $p(\theta \mid Y)$
`r vspace("20px")`
.box-content-green[
It normalises the posterior probabilities so they sum to 1
]

- So we can simply acknowledge that  $p(y=10) = 1/c$ and replace the posterior as:

$$p(\theta \mid y=10) \propto p(\theta)\times L(\theta \mid y)$$
- The proportionality means that if we compare the normalised and unnormalised posterior they preserve their relative relationship
]

.pull-right[
```{r eval=TRUE, fig.height = 10, fig.width = 15, fig.show='hold',opts=list(width="110%")}
df_nn <- df[,2]*df[,4]
norm <- df_nn/sum(df_nn)
p1<-ggplot(df, aes(df[,1],df_nn)) + geom_point()  + geom_segment(aes(x=df[,1],y=0,xend=df[,1],yend=df_nn)) + xlab(expression(theta)) + ylab(expression(paste("p(",theta, "| y=10)"))) + ggtitle("Posterior non normalised") + ylim(0,0.1) + 
    theme(axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold", hjust=0.5))


p2<-ggplot(df, aes(df[,1],norm)) + geom_point()  + geom_segment(aes(x=df[,1],y=0,xend=df[,1],yend=norm)) + xlab(expression(theta)) + ylab(expression(paste("p(",theta, "| y=10)"))) + ggtitle("Posterior normalised") + ylim(0,1) +   
  theme(axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold", hjust=0.5))
grid.arrange(p1,p2,nrow=1)
```
]
---

name: recap

`r vspace("250px")`

.myblue[.center[.huge[
**A quick recap**]]]


---




# So a quick general recap: Bayesian inference

Makes fundamental distinction between 
- Observable quantities $y$, i.e.~the data
`r vspace("20px")`
- Unknown quantities $\theta$

`r vspace("30px")`

  - $\theta$ can be statistical parameters, missing data, mismeasured data...
`r vspace("20px")`
$\rightarrow$ parameters are treated as random variables
`r vspace("20px")`
$\rightarrow$ in the Bayesian framework, we make probability statements about model parameters

--

`r vspace("50px")`

.red[Note that in the Frequentist framework, parameters are fixed non-random quantities and the probability statements concern the data]

---

# Bayesian inference

- As with any statistical analysis, we start building a model which specifies $p(Y=y \mid \theta)$
`r vspace("20px")`

- This is the .red[data distribution], which relates all variables into a .red[**full probability model**]
`r vspace("20px")`

- The choice of data distribution depends on the nature of the data:

e.g. are we analysing continuous or discrete data, are the data symmetric or skewed, etc.
`r vspace("20px")`

- As we observe the data, we can use descriptive tools (e.g. plots) to visualise the data and choose the best likelihood
`r vspace("20px")`

---

# Bayesian inference

From a Bayesian point of view
`r vspace("20px")`

-  $\theta$ is unknown so should have a .red[probability distribution] reflecting our uncertainty about it before seeing the data
     
`r vspace("20px")`

$\rightarrow$ need to specify a .redp[prior distribution] $p(\theta)$
`r vspace("20px")`

- $y$ is known so we should condition on it
`r vspace("20px")`
$\rightarrow$ use Bayes theorem to obtain conditional probability distribution for unobserved quantities of interest given the data:

$$p(\theta \mid y)= \frac{ p(\theta)\, p(y \mid \theta)}
                         {\int  p(\theta)\,p(y \mid \theta)\,d\theta}
	\propto p(\theta)\,p(y \mid \theta)$$
`r vspace("20px")`

This is the .red[posterior distribution]

--

- The prior distribution $p(\theta)$, expresses our uncertainty about $\theta$ .red[before] seeing the data

- The posterior distribution $p(\theta \mid y)$, expresses our uncertainty about $\theta$ .red[after] seeing the data


---

# The posterior distribution

Posterior distribution forms basis for all inference --- can
be summarised to provide
`r vspace("10px")`
- point and interval estimates of Quantities of Interest (QOI), e.g. treatment effect, small area estimates, ...
`r vspace("10px")`
- point and interval estimates of any function of the parameters
`r vspace("10px")`
- probability that QOI (e.g. treatment effect) exceeds a critical threshold
`r vspace("10px")`
- prediction of QOI in a new unit
`r vspace("10px")`
- prior information for future experiments, trials, surveys, ...
`r vspace("10px")`
- inputs for decision making
`r vspace("10px")`
- ...

---

name: Conjugacy

`r vspace("250px")`

.myblue[.center[.huge[
**What is conjugacy?**]]]


---

# How to select a prior

- Selecting the prior is crucial for a Bayesian analysis

.content-box-green[
- There is no right way to select a prior
- The choices often depend on the objective of the study and the nature of the data 
`r vspace("-10px")`
]

- There are other criteria to consider when choosing a prior model:

**Computational ease**

Especially if we don’t have access to computing power, it is helpful if the posterior model is easy to build.

**Interpretability**

The posterior balance (is a compromise) the data and the prior. A posterior model is interpretable, and thus more useful, when you can look at its formulation and identify the contribution of the data relative to that of the prior.

--

We introduce now a type of models which satisfy both properties

.content-box-yellow[
`r vspace("-20px")`
### Conjugate prior
`r vspace("-20px")`
Let the prior model for a parameter $\theta$ with $p(\theta)$ and the model of data $Y$ conditioned on $\theta$ have likelihood function $L(\theta \mid y)$.

If the resulting posterior model $p(\theta \mid y) \propto p(\theta) \times p(y \mid \theta)$ is of the same family as the prior, then we say it is a **conjugate prior**.
`r vspace("-10px")`
]


---

name: Beta-Binomial

`r vspace("250px")`

.myblue[.center[.huge[
**Some conjugacy models: Beta-Binomial**]]]

---


# Beta-Binomial model for proportions: example

- We consider an early investigation of a new drug
`r vspace("30px")`
- Experience with similar compounds has suggested that response rates between 0.2 and 0.6 could be feasible
`r vspace("30px")`
- We interpret this as a distribution with mean = 0.4 and standard deviation 0.1 
`r vspace("30px")`
- A Beta(9.2,13.8) distribution has these properties (Check recording 7 to see how to go from the .red[mean and sd] to the .red[a and b] parameters of a Beta distribution)
`r vspace("30px")`
- Suppose we now treat $n=20$ volunteers with the compound and observe $y=15$ positive responses

---

# Identifying the different model components

- Assuming patients are independent, with common unknown response rate $\theta$, leads to a binomial data distribution
\begin{align}
p(y \mid  n, \theta) &= \left( \begin{array}{c} n \\ y \end{array} \right) \theta^y (1-\theta)^{n-y} \; \propto \; \theta^y (1-\theta)^{n-y}
\end{align}


- $\theta$  needs a continuous prior distribution:

\begin{align}
\theta & \sim  \hbox{Beta}(a,b )\\
p(\theta) &= \frac{\Gamma (a+b)}{\Gamma (a) \Gamma(b)} \; \theta^{a-1} (1-\theta)^{b-1}
\end{align}

---

# Combining prior and data

Combining the Binomial data and the Beta prior gives the following posterior distribution

.content-box-green[
\begin{align}
p(\theta \mid y, n)  & \propto  p(y \mid \theta, n) p(\theta)\\
& \propto  \theta^y (1-\theta)^{n-y} \theta^{a-1} (1-\theta)^{b-1}\\
& =  \theta^{y+a-1} (1-\theta)^{n-y+b-1}
\end{align}
]

The posterior is still a Beta distribution (with different parameters):

$$p(\theta \mid y, n)  \propto  \hbox{Beta}(y+a, \; n-y+b)$$
---

# Comparing prior, likelihood and posterior

```{r eval=TRUE, fig.height = 12, fig.width = 10, fig.show='hold',opts=list(width="40%")}
#Prior parameters
a=9.2;b=13.8
#Data
y=15; n=20

library(cowplot)
p1 <- ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(fun = dbeta, n = 101, args = list(shape1 = a, shape2 = b)) + ylab(expression(p(theta))) + scale_y_continuous(breaks = NULL)
p_temp <- dbinom(15,20,seq(0,1,0.01))
x<-seq(0,1,0.01)
p2 <- ggplot() + aes(seq(0, 1,0.01), p_temp)+geom_line() + ylab(expression(paste("L(",theta,"|y=15)"))) + scale_y_continuous(breaks = NULL) + xlab("x")
p3 <- ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(fun = dbeta, n = 101, args = list(shape1 = a+15, shape2 = b+5)) + ylab(expression(paste("p(",theta,"|y=15)"))) + scale_y_continuous(breaks = NULL)
grid.arrange(p1,p2,p3,nrow=3)
```

---

# Gamma-Poisson model for count data: example

- .red[For a recap on the Poisson distribution see recording 5]

`r vspace("20px")`

- In epidemiology we are often interested in estimating the .red[rate] or .red[relative risk] rather than the .red[mean] for Poisson data:

`r vspace("20px")`

  - Suppose we observe $y=7$ cases of leukaemia in one region;
  - The expected number of cases is $E=4$

`r vspace("20px")`

  - **Data distribution**: Poisson with mean $\theta = \lambda \times E$, where $\lambda$ is the unknown incidence ratio:
  
$$p(y \mid  \lambda, E) = \frac{(\lambda E)^{y}e^{-\lambda E} }{y!}$$
`r vspace("20px")`

  - **Prior**: $\mathrm{Gamma}(a, b)$ on the the risk $\lambda$:
$$p(\lambda)  =  \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b \lambda} $$

.red[Check recording 8 for a recap on the Gamma distribution]
---

# Combining likelihood and prior

This implies the following posterior

\begin{align*}
p(\lambda \mid y) & \propto p(\lambda)\,p(y \mid \lambda) \\
  & = \frac{b^a}{\Gamma(a)}\lambda^{a-1}e^{-b(\lambda)}e^{-(\lambda E)}\frac{(\lambda E)^{y}}{y!}\\
  & \propto  \lambda^{a+y-1} e^{-(b+E)\lambda}\\
 & =  \mathrm{Gamma}(a + y ,\, b + E)
\end{align*}

--

The posterior is another (different) Gamma distribution

$$E(\theta \mid y)= \frac{a + y }{ b + E}$$

So posterior mean depends on the prior $(a, b)$ and on the data $(y, E)$


---

# Prior, likelihood, posterior for the Leukaemia example

.pull-left[

- Assuming a prior $\lambda \sim \mathrm{Gamma}(2,2)$
`r vspace("20px")`

- Considering the data $y=7, E=4$ 
`r vspace("20px")`

- We obtain a posterior $\lambda \mid y \sim \mathrm{Gamma}(9,6)$ centered around 1.5

]

.pull-right[

`r include_fig("PoissonGamma.png", width="75%")`

The posterior becomes a compromise between the prior and the data
]

---

# References

```{r refs, echo=FALSE, results="asis"}
PrintBibliography(bibfile,.opts=list(max.names=3))
```
