---
title: "Practical 6b - Disease mapping study of COVID-19 mortality in England, March-July 2020"
author: "Bayesian modelling for spatial and spatio-temporal data"
output: 
  html_document:
    toc: true
    toc_float: true
bibliography: biblio.bib
editor_options: 
  chunk_output_type: console
---

\pagenumbering{gobble} 
\pagenumbering{arabic} 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.align = "center", class.source='klippy')
```
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'),color = 'darkred',
               tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

In this lab we will carry out a disease mapping study (without spatial structure) of COVID-19 mortality in England during March-July 2020, partitioned in lower tier local authorities (LTLAs) and evaluate the impact of the deprivation to explain the spatial variation.

We will also work with different R packages for the visualization of vector data, and we will learn how to produce effective visualizations of our data and results.


* To start with, create a separate subdirectory to save your files created during this tutorial (e.g. setwd("~/Labs/Lab2b")).

# 1. Install and load packages 

This Lab requires the following packages to be installed and attached: `dplyr`, `sf`, `spdep`, `RColorBrewer`, `ggplot2`, `tmap`, `cowplot`, `mapview`. 

* To check whether a package is installed you can use the query:
```{r echo=TRUE, eval=FALSE}
is.element("sf", installed.packages())
```

* If FALSE is returned, then install the package function `install.packages()`, e.g.
```{r eval = FALSE,  results="hide"}
install.packages("sf", dep=TRUE)
```

* To install the entire suite of packages, we can use:
```{r eval = FALSE,  results="hide"}
#install.packages(c("dplyr","sf","spdep","RColorBrewer", "ggplot2", "tmap", "mapview", "cowplot"), dependencies = TRUE, repos = "http://cran.r-project.org")
```

* Then, load the needed packages:
```{r eval = TRUE, results="hide", message=FALSE, warning=FALSE}
library(INLA)         # Integrated Nested Laplace Approximation package
library(dplyr)        # A package for data manipulation
library(sf)           # Simple feature for R
library(spdep)        # Functions and tests for evaluating spatial patterns 
                      # and autocorrelation

# Packages used for visualization 
library(RColorBrewer) # A package providing colour palettes for shading maps 
                      # and other plots
library(tmap)         # A package for static and interactive maps
library(ggplot2)      # A package that implements the grammar of graphics, which is a term used to
                      # break up graphs into semantic components, such as geometries and layers.
library(mapview)      # A package for interactive maps
library(cowplot)      # Add-on to ggplot. It provides features that help us with creating
                      # publication-quality figures

```
NOTE: if you have issues in working with `mapview` package in R Markdown, try to install it with `remotes::install_github("r-spatial/mapview")`.

# 2. Visualization of spatial areal data

We start by reading the shapefile of England at the LTLA level (317 areas) using the function `st_read()` of the package `sf`. Call this object as LTLA:
```{r eval=TRUE}
LTLA = st_read("LTLA_shp.shp")

class(LTLA)
```


* Now, we examine the sf object London. 
The key metadata for the shapefiles include:

1. Object Type: the class of the imported object.
2. Coordinate Reference System (CRS): the projection of the data.
3. Extent: the spatial extent of the shapefile (that is the geographic area that the shapefile covers). 

```{r eval=FALSE}
# Check geometry
st_geometry_type(LTLA) 

# Check what CRS this file data is in
st_crs(LTLA) # the data are encoded using a Transverse Mercator Projection. 
               # The Airy ellipsoid is being used (+ellps=airy) and the units are meters (+units=m)

# Check the spatial extent of the shapefile (i.e. the geographic "edge" or location that is the furthest north, south east and west) 
st_bbox(LTLA)

# View all of the metadata and attributes for this shapefile object
LTLA
```

We can create basic maps of sf objects using the `plot()` function:

```{r eval=FALSE}
# the default plot of an sf object is a multi-plot of all attributes

plot(LTLA)          # plot all the attributes
plot(LTLA$geometry) # plot only the boundaries
```

Note: A particular CRS can be referenced by its EPSG code. The EPSG is a structured dataset of CRS and Coordinate Transformations. It was originally compiled by the European Petroleum Survey Group (now defunct).

Now, we obtain static and interactive maps.

## Static map with `ggplot2` 

The package `ggplot2` implements the *grammar of graphics* [@Wilkinson_2005] in R, which is a term used to break up graphs into semantic components, such as geometries and layers.
Spatial objects can be plotted directly with `ggplot2` using the `geom_sf()` function, without converting it into a dataframe. In particular, the `geom_sf` function adds a geometry stored in a `sf` object. 
Plotting layers can be added with the `+` operator.

```{r eval=TRUE, fig.cap =""}
ggplot() + 
      geom_sf(data = LTLA, color = "red", fill = "white") + 
      ggtitle("Map of LTLAs in England") + 
      coord_sf() +    #axis limits and CRS
      labs(x = "Longitude", y = "Latitude", fill = "") +
      theme_bw() +    # dark-on-light theme
      theme(axis.title = element_text(size = 16),
            axis.text = element_text(size = 14))
```


## Static map with `tmap`

Like `ggplot2`, `tmap` is based on the idea of a *grammar of graphics* . This involves a separation between the input data and the aesthetics (that is, how data are visualised): data can be mapped in a range of different ways including location on the map (defined by data's geometry), color, and other visual functions. 
The basic building block is `tm_shape()` (which defines input data, raster and vector objects), followed by one or more layer elements such as `tm_fill()` etc. (To start with, you may want to visit the page: https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

```{r LTLA_map, eval=TRUE}
tm_shape(LTLA) +
          tm_fill("tomato") +
          tm_borders(lty="dashed", col="gold") +
          tm_style("natural", bg.color="grey90") +
          tm_layout(title="Map of LTLAs in England")
```

NOTE: Here `tmap` style is set to "natural". Other available styles are: "white", "gray", "natural", "cobalt", "col_blind", "albatross", "beaver", "bw", "watercolor".


## Interactive maps with `tmap`
We previously saw a static map made with `tmap`. However, `tmap` allows to plot not only a static image but also an interactive image. Using `tmap_mode()`, we can switch between presentation (`"plot"`) and interactive (`"view"`) modes. For example:

```{r eval=TRUE, message=FALSE}
# as seen before, but with specified presentation mode
tmap_mode("plot")
tm_shape(LTLA) +
            tm_fill("grey90") +
            tm_borders(lty="solid", col="skyblue1") +
            tm_style("albatross") +
            tm_layout(title="Static Map of LTLAs in England")
```

```{r eval=TRUE, message=FALSE}
# interactive mode
tmap_mode("view")
tm_shape(LTLA) +
            tm_fill("grey90") +
            tm_borders(lty="solid", col="black") +
            tm_style("natural") +
            tm_layout(title="Interactive Map of LTLAs in England")
```


## Interactive map with `mapview` (the computation is slightly demanding)
We can also produce a web map visualization as follows:
```{r eval=FALSE, echo=TRUE}
mapview(LTLA, legend = FALSE)
```


# 3. Import and explore the data

The data set includes the number of COVID-19 deaths and expected counts for the 317 Local Authorities in England. In addition, it includes the value of the Index for Multiple Deprivation (IMD), which is a complex score measuring the deprivation of the area across several domains, includind socio-economic, crime, environment, healthcare. 

1. Import the health data
```{r eval=TRUE, echo = FALSE}
COVID19Deaths = read.csv("COVID19Deaths.csv", header = TRUE)
```

2. Calculate the SMRs and add them to the data frame COVID19Deaths
```{r eval=TRUE, echo = TRUE}
COVID19Deaths$SMR = COVID19Deaths$O/COVID19Deaths$E
```

To produce maps we need to join the health data with the `sf` object. To do so, we take advantage of shared key variables.
We use the library `dplyr` that has multiple join functions. `dplyr` join functions work the same on data frames and `sf` objects, the only important difference being the `geometry` list column.
The most common type of attribute join on spatial data takes an `sf` object as the first argument and adds columns to it from a `data.frame` specified as the second argument.

```{r eval=TRUE}
names(LTLA)
names(COVID19Deaths)

England_SMR = left_join(LTLA, COVID19Deaths, by = c("LTLA" = "LTLA"))
```
Note: Because the input datasets share the 'key variable' (LTLA) the join would have also worked without using the `by` argument.

We can now display maps with all variables of England_SMR.

* We can quickly display number of deaths
```{r eval=TRUE}
plot(England_SMR["O"])
```

* Then, we can display on a map the SMRs
```{r eval=TRUE}
plot(England_SMR["SMR"], breaks = c(0,0.5,1,1.5,2))
```

* We can use`ggplot2` and `tmap` to create a map of the SMRs.

+ With `ggplot2` we can obtain a simple map as follows:
```{r eval=TRUE, echo=TRUE, eval=TRUE}
MapSMR1 = ggplot() + 
  geom_sf(data = England_SMR, aes(fill = SMR), col = NA) +
  scale_fill_viridis_c(limits = c(0,2)) + theme_bw()  

MapSMR1
```


+ and with `tmap` as follows:
```{r SMR_map, eval=TRUE}
tmap_mode("plot")
tm_shape(England_SMR) + tm_polygons("SMR",palette="RdYlGn", style="cont", n=8) +
            tm_borders(lty="solid", col="black") +
            tm_style("natural") +
            tm_layout(title="Map of SMRs at LTLA level in England")
```
As we did with `ggplot2`, we can create a map object with `tmap`.


+ We can also map the IMD as follows:
```{r IMD_map, eval=TRUE}
tmap_mode("plot")
tm_shape(England_SMR) + tm_polygons("IMD",style="pretty", n=8, alpha=0.5) +
            tm_borders(lty="solid", col="black") +
            tm_style("natural") +
            tm_layout(title="Map of IMD at LTLA level in England")
```

A useful package that can be used when mapping spatial objects is `RColorBrewer`. Once loaded, you can see a list of all the color pallets typing `display.brewer.all()` (see also Lab2a for details on this package), then pick a palette you like and plot the SMRs
```{r eval=TRUE}
display.brewer.all()
```

Thus, the palettes from `RColorBrewer`can be used in `ggplot2`, e.g.: 
```{r eval=TRUE}
breaks =  c(0,0.5,1,1.5,2)
England_SMR = mutate(England_SMR, SMR_cat = cut(SMR, breaks, include.lowest = TRUE)) # mutate() adds new variables and preserves existing ones

ggplot() + geom_sf(data = England_SMR, aes(fill = SMR_cat), col = NA) +
  theme_bw() + scale_fill_brewer(palette = "OrRd") + 
  guides(fill=guide_legend(title="SMR"))

```


# 4. Obtaining the posterior relative risks for the disease mapping (RRs)

The RRs will be smoothed using the Poisson-logNormal hierarchical model. 

We specify the model using a `iid` prior and then run in in `INLA`. The `iid` latent effect, has only has the precision hyperparameter. We can inspect the internal representation (`theta`), the name of the parameter (`log precision`) and a short name (`prec`): 

```{r, eval=TRUE}
names(inla.models()$latent$iid$hyper)
```

```{r, eval=TRUE}
inla.models()$latent$iid$hyper$theta$name
```

```{r, eval=TRUE}
inla.models()$latent$iid$hyper$theta$short.name
```

In order to define a prior, both the name of the internal parameter `theta` and the short name `prec` can be used in the named list passed to hyper. However, regardless of the name used, the prior is always set in the scale of the internal parameter, i.e., if the prior is set using prec it must be a prior on the log-precision (see for details Virgilio Gomez-Rubio's book " Bayesian inference with INLA" https://becarioprecario.bitbucket.io/inla-gitbook/ch-priors.html)


Now, run the model
```{r INLA_iid, eval=TRUE}
ID = seq(1,317)

# we define a Gamma prior on the precision of a iid model with parameters  
# 0.01 and 0.01
prec.prior = list(prec = list(prior = "loggamma", param = c(0.01, 0.01))) 

formula_iid = O ~ f(ID, model="iid", hyper = prec.prior)	

mod_iid = inla(formula=formula_iid, family="poisson", data=England_SMR, E=E, control.compute=list(dic=TRUE, waic=TRUE))
```

* To see the DIC and WAIC 
```{r eval=TRUE}  
mod_iid$waic$waic
mod_iid$dic$dic
```	

# 5. Obtaining the posterior relative risks for the ecological regression (RRs)

* Now let's add the IMD as a covariate in the model
```{r INLA_reg, eval=TRUE}
formula_reg = O ~ IMD + f(ID, model="iid", hyper = prec.prior)	

mod_reg = inla(formula=formula_reg, family="poisson", data=England_SMR, E=E, control.compute=list(dic=TRUE, waic=TRUE))

mod_reg$waic$waic
mod_reg$dic$dic
```
Inspecting the WAIC and DIC, we can conclude that the two models are very similar in terms of fitting, so we will focus on the simpler hierarchical model in the maps.


# 6. Map of the smoothed RRs

* To map the smoothed RRs in R we extract the posterior mean of the relative risks (RRs). Note that we take the mean on the transformed marginals to get the log-RR back to the original scale:
```{r eval=TRUE} 
RR_COVID = c()
for(i in 1:317){
  RR_COVID[i] = inla.emarginal(function(x) exp(x), 
        mod_iid$marginals.random$ID[[i]])
}

```	

* Add it on the shapefile
```{r eval=TRUE} 
England_SMR$RR = RR_COVID
```	

* Using `ggplot2`, we can produce a map of the smoothed RRs
```{r eval=TRUE} 
MapRR1 = ggplot() + geom_sf(data = England_SMR, aes(fill = RR), col = NA) + theme_bw() + 
                     scale_fill_viridis_c(limits = c(0,2)) 

MapRR1
```	

* Now we have our maps sorted, we can arrange them side by side in the same plot using the `cowplot` package, which is a simple add-on to `ggplot`. Comment the maps.
```{r eval=TRUE} 
plot_grid(MapSMR1, MapRR1, ncol = 1, align = 'v', labels="AUTO", rel_widths = c(1, 1))
```	

NOTE: we can specify labels of figures as `labels = "AUTO"` or `labels = "auto"` then labels will be auto-generated in upper or lower case, respectively.


+ Exporting maps with `ggplot2` (examples)
```{r eval=TRUE}
## save an image - ggplot2
ggsave("MapRR1.jpeg", plot = MapRR1) # if plot is not specified it saves the last plot.
```
