---
title: "Spatio-temporal Bayesian models for environmental epidemiology"
author: Marta Blangiardo and Garyfallos Konstantinoudis
institute: "[Environment and Health Statistics group](https://www.envstats.org/) | Imperial College London"
params: 
   conference: "MRC Health Collaboration with Helmholtz Centres"
   location: "London"
   date: 16 May 2024
   short_title: "Spatio-temporal Bayesian models for environmental epidemiology"
output:    
  xaringan::moon_reader: 
    includes: 
       in_header: "assets/latex_macros.html" 
       # This line adds a logo based on the format selected in the file 'assets/include_logo.html'
       # NB: the actual options (eg placement of the logo and actual logo file) can be changed there
      # after_body: "assets/insert-logo.html"
    seal: false
    yolo: no
    lib_dir: libs
    nature:
      beforeInit: ["https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
      titleSlideClass:
      - center
      - middle
    self_contained: false 
    css:
    - "assets/beamer.css"

---

```{r global_options, echo = FALSE, include = FALSE}
options(width = 999)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
```

```{r echo=F,message=FALSE,warning=FALSE,comment=NA}
# Sources the R file with all the relevant setup and commands
source("assets/setup.R")
library(INLA)

# Stuff from 'xaringanExtra' (https://pkg.garrickadenbuie.com/xaringanExtra)
# This allows the use of panels (from 'xaringanExtra')
xaringanExtra::use_panelset()
# This allows to copy code from the slides directly
#xaringanExtra::use_clipboard()
# This freezes the frame for when there's a gif included
#xaringanExtra::use_freezeframe()

# Defines the path to the file with the .bib entries (in case there are references)
bibfile=RefManageR::ReadBib("refs.bib",check = FALSE)
```
class: title-slide

# `r rmarkdown::metadata$title``r vspace("10px")` `r rmarkdown::metadata$subtitle`

## `r rmarkdown::metadata$author`

### `r rmarkdown::metadata$institute`    

### `r rmarkdown::metadata$params$conference`, `r rmarkdown::metadata$params$location` 

<!-- Can also separate the various components of the extra argument 'params', eg as in 
### `r paste(rmarkdown::metadata$params, collapse=", ")`
-->

`r ifelse(is.null(rmarkdown::metadata$params$date),format(Sys.Date(),"%e %B %Y"),rmarkdown::metadata$params$date)`

`r vspace("20px")`

`r include_fig("MRC-Centre-Logo.png", width="30%")`

---

layout: true  

.my-footer[ 
.alignleft[ 
&nbsp; &copy; Blangiardo & Konstantinoudis
]
.aligncenter[
`r rmarkdown::metadata$params$short_title` 
]
.alignright[
`r rmarkdown::metadata$params$conference`, `r short_date` 
]
] 

```{css,echo=FALSE, eval=FALSE}
.red {
  color: red;
}
.blue {
  color: 0.14 0.34 0.55;
}

.content-box-blue { background-color: #F0F8FF; }

}
```
<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>


---

# Data/modelling/challenges in environmental epidemiology

`r include_fig("Areas.png",width="80%")`

---

# Accounting for spatial and temporal dependencies

.center[.content-box-blue[
**First law of geography** (W. Tobler, 1970): everything is related to everything else, but near things are more related than distant things
]
]

`r vspace("20px")`

When are we interested in the spatial and temporal components?

--

`r vspace("20px")`

- Hypothesis generating perspective 

$\rightarrow$ Are we explicitly interested in the spatial pattern of disease risk?

$\rightarrow$ Do we want to evaluate temporal trends for each area?

--

`r vspace("20px")`

- Is the spatial clustering and/or temporal trend a nuisance quantity that we wish to take into account but are not explicitly interested in?

$\rightarrow$ Spatial / time series regression.


--

`r vspace("40px")`

In this talk we are going to focus on

`r vspace("10px")`
  - Wastewater modelling and COVID-19 prevalence
`r vspace("10px")`
  - Temperature and COPD hospitalisations

---

name: WBE

`r vspace("250px")`

.myblue[.center[.huge[
**Space-time modelling: wastewater based epidemiology**]]]

---

#Wastewater-based epidemiology

`r vspace("50px")`
<center>
.content-box-green[.center[.large[A collection of tools
and methods for surveillance and monitoring disease outbreaks using bio-
chemical analysis of wastewater samples as the primary outcome measure]]]
</center>

--

`r vspace("20px")`

- Wastewater has been used over the years as an epidemiological tool
  - no selection bias
  - cheap

- Methods to detect the presence of SARS-CoV-2 RNA in wastewater have
been developed in a number of countries `r Cite(bibfile,"tlhagale2022establishment")`

- Scientific contributions to date have focused on the level of the sewage treatment works (STWs) where the
measurements of SARS-CoV-2 RNA from wastewater are obtained.

- No attempt has previously been made to develop a
spatially resolved model to estimate wastewater viral
concentration over a spatially continuous domain.

---

# How does it work 

`r vspace("10px")`
`r include_fig("WW.png",width="80%", title="")`

### Aim

- Develop a statistical model to estimate viral concentration in wastewater at fine spatio-temporal resolution covering the whole of England
`r vspace("10px")`
- Use the spatial dependence to **predict** values of the spatial surface (together with associated uncertainty) at locations where there are no observations.


---

# Data

- SARS-CoV-2 viral concentrations obtained from the 303 Sewage Treatment Works (STWs) for which the data were publicly available from the Environmental Monitoring for Health Protection (EMHP) wastewater surveillance programme.

.pull-left[
`r include_fig("STW.png", width="70%", title="Location of STWs and their catchment areas")`
.center[Location of STWs and their catchment areas]
]

.pull-right[
- Socio-demographic, land use covariates and wastewater genomic data:
  1. Index of Multiple Deprivation (2019)
  2. Black, Asian and Minority Ethnic (BAME) proportion (Census, 2011)
  3. Land cover (Copernicus, 2018)
  4. Population density (ONS, 2019)
  5. Age structure (ONS, 2019)
  6. Percentage coverage of the SARS-CoV-2 genome in the sample
  7. Single-Nucleotide Polymorphism (SNP) data in the sample
]

---

# Statistical model

The log-transformed number of gene copies per liter at catchment area $i=1, \ldots, 303$ and week $t=1,\ldots, 44$ is modelled as:

$$y_{it} \sim \text{Normal}(\mu_{it}, \sigma^2_y)$$

### Geospatial model on the latent mean concentration at monitoring site as a function of:
1. covariates
2. regional effect
3. temporal trend
4. spatio-temporal random effect

--

`r vspace("20px")`

### Spatial prediction

1. at the population-weighted centroids of all the 32,844 Lower Super Output Areas (LSOAs) in England 
2. Aggregated at the Lower Tier Local Authority level 
3. The posterior predictive distribution is used to detect increases in viral concentration over two consecutive weeks

--

`r vspace("20px")`

### Bayesian approach implemented in R-INLA 

---


#Interactive dashboard
To visualise these rich model outputs, we have created a dynamic and interactive dashboard

`r vspace("-50px")`

```{r, echo=FALSE, out.width="140%", out.extra='style="transform: scale(.8);border:none"'}
options(htmltools.dir.version = FALSE, htmltools.preserve.raw = FALSE)
knitr::include_url("https://b-rowlingson.gitlab.io/wwatlas/", height="100%")
```

---

# Combining wastewater viral concentration and COVID-19 prevalence

We consider the use of wastewater data under the following
setting:
`r include_fig("WWCovid.png", width="70%")`

`r vspace("20px")`

###Aim 

To **integrate** the spatially granular wastewater data with prevalence
estimates in order to downscale prevalence estimates from
the national to a local level when only the national prevalence is
available.


---

# Modelling framework

- Coarse (national) and local (LTLA) models jointly fitted

- Wastewater predictions available from the geostatistical model

- Posterior distribution of COVID-19 prevalence obtained through Nicholson et al. (2022) 

--

.pull-left[
### Coarse (national)
\begin{align*}
\mbox{logit}({P}_{t}) &\sim \mbox{N}(\mu_{t},\sigma^2_{t})\\
\mu_{t} &= \alpha + b_t + \delta_t \tilde{y}^{\prime}_{t}
\end{align*}

to learn about
1. the time pattern of prevalence: $b_{1:T}$
2. the national level WW - prevalence relationship: $\delta_t \tilde{y}^{\prime}_{t}$ 

Both are modelled using autoregressive structures
]

.pull-right[
### Local
\begin{align*}
\mbox{logit}(p_{lt}) &\sim \mbox{N}(\mu_{lt},\sigma^2_{lt})\\
\mu_{lt} &= (\alpha + u_l) + (b_t + v_{lt}) + (d_t + m_l)y^{\prime}_{lt}\\
\end{align*}

to learn about
1. the local departure from the national trend of prevalence $v_{lt}\sim N(0,\sigma^2_v)$
2. the local departure from the  WW-prevalence relationship $m_l \sim N(0,\sigma^2_m)$
 ]

`r vspace("20px")`

.small[Nicholson, G, et al. "Improving local prevalence estimates of SARS-CoV-2 infections using a causal debiasing framework." Nature Microbiology 7.1 (2022): 97-107.]

---

# Added values of WW: more accurate, less uncertain nowcasts

.pull-left[`r include_fig("Nowcast.jpg", width="80%")`
]

.pull-right[
- Using wastewater data helps to improve the quality of the disaggregated LTLA prevalence (lower RMSE and better coverage).

`r vspace("30px")`
`r include_fig("RMSE.jpg", width="120%")`
]

---

# Take home message

- The proposed framework for disaggregation demonstrates the added-value of wastewater data to inform local prevalence of COVID

`r vspace("20px")`

- Example of .red[modular] integration of environmental and disease data  



.pull-left[
`r vspace("20px")`

- Currently being extended to include other data sources, such as syndromic and secondary care data, to enable quantitatively accurate monitoring of disease evolution across a population.
]

.pull-right[
`r include_fig("Surveillance_representation.jpg", width="80%")`
]


---

name: COPD

`r vspace("250px")`

.myblue[.center[.huge[
**Space-time modelling: Temperature and Health**]]]



---

# COPD and temperature
.panelset[
.panel[.panel-name[COPD]

.pull-left[
- 3rd cause of death, 3.17 million deaths in 2015 globally
`r vspace("20px")`
- In England, 115,000 emergency admissions and 24,000 deaths per year.
`r vspace("20px")`
- COPD exacerbations: Bacteria, viruses and air-pollution.
`r vspace("20px")`
- The role of temperature is unclear. 
]

.pull-right[

`r include_fig("COPD_HES.jpg", width="80%")`
]

]

.panel[.panel-name[Temperature and Health]

.pull-left[
- Typically U-shaped relationship between temperature and mortality.
`r vspace("20px")`
- Cold, dry air or hot air can trigger a flare-up.
`r vspace("20px")`
- Different confounding, different lags across different temperatures.
`r vspace("20px")`
- This study focuses on warm temperatures.

]

.pull-right[

`r include_fig("TemperatureHealth.png")`
]

]
]

---

# Previous studies

```{r lit-table}
library(kableExtra)
library(dplyr)

Authors <- c("Michelozzi et al 2009","Anderson et al 2013","Zhapo et al 2019")
Aggregation <- c("city and daily","county and daily","individual")
Country <- c("EU","US","Brazil")
Pollutants <- c("NO2, O3", "O3, PM10, PM2.5", "no adjustment")
Effect <- c("2.1 (0.6 to 3.6) per 1oC", "2.0 (0.4 to 4.5) per 10oF", "5.0 (4.0 to 6.0) per 5oC")


df<- data.frame(Authors, Aggregation, Country, Pollutants, Effect)
df %>% 
  kbl() %>% 
  kable_material(c("striped", "hover"))
```


- Spatial and temporal aggregation
  - Exposure varies on high resolution
  - Insufficient adjustment for confounding
  - Ecological bias.
`r vspace("20px")`
- One study individual data, but did not account for air-pollution

---

# Aim

`r vspace("30px")`
.center[.content-box-green[Examine the effect of warm temperatures on COPD hospitalizations.]]

`r vspace("30px")`

Secondary objectives:
`r vspace("20px")`
- Examine effect modification by age, sex and space.
`r vspace("20px")`
- Assess how contextual characteristics, such as green space, average temperature, deprivation and urbanicity affect population vulnerability. 
`r vspace("20px")`
- Calculate the number of COPD cases attributable to temperature.

---

# Outcome and exposure

.pull-left[
### Outcome
- COPD hospitalization (ICD10 J40-44) 2007-2018 (NHS digital).
- Individual data/ 100m grid spatial resolution.
- Summer months.

### Exposure
- Daily maximum temperature 2007-2018 at 1km grid from MetOffice.
- lag0-2.
]

.pull-right[
`r include_fig("temperatureLTLA.png", width="80%")`
]

---

# Case-crossover design

- Epidemiological study design for transient environmental exposures
`r vspace("20px")`    
- Sample from same month and days
`r vspace("20px")`    
- Exposure period 0-2 days before the hospital episode (mean temperature)

`r include_fig("FigCaseCrossOver.png", width="70%")`

- Adding a spatial dimension to account for modifiers
---

# Confouders and spatial effect modifiers

.panelset[
.panel[.panel-name[Confounders]

```{r conf-tables}
library(kableExtra)
Confounder<- c("PM2.5","O3","Relative Humidity","Holidays")
Source <- c("MetOffice","MetOffice","MetOffice","ONS")
Space<-c("1Km","1Km","10Km", "nationwide")
Time<- rep("daily",4)
Year<-rep("2007-2018",4)

df<- data.frame(Confounder,Source,Space,Time,Year)
df %>% 
  kbl() %>% 
  kable_material(c("striped", "hover"))
```
]

.panel[.panel-name[Effect modifiers]

`r include_fig("SpatEffMod.png", width="70%")`
]
]

---

# Step 1: Find the linear threshold

Let $Y_{tj}$ be the case-control identifier for the COPD hospitalisation for the event (case or control) at time $t$, in the $j-th$ case-control group:
  
\begin{align}
Y_{tj} &\sim \mbox{Poisson}(\mu_{tj})\\
\log(\mu_{tj}) &= \alpha_1 I(X_{1t} < c) X_{1t} + \alpha_2 I(X_{1t} \geq c) X_{1t} + \sum_{m=1}^4\beta_mZ_{mt} + u_{j}\\
u_j &\sim N(0, 100)\\
\alpha_1, \alpha_2, \beta_1, \dots \beta_4 & \sim N(0, 1)\\
\end{align}

$c$ is the temperature threshold,  $50-$th, $55-$th, ...,  $90-$th,  $95-$th percentile. 

--

`r vspace("20px")`

We ran the above model for the different temperature thresholds and chose the one which minimises the WAIC.

---

# Step 2a: Effect modification by age and sex
  
Similarly we fitted the previous model for the different sex-age groups (male/female and $<65$, $65-85$, $>85$ ). Consider $g$ the index for the age-sex groups:

\begin{align*}
Y_{tjg} &\sim \text{Poisson}(\mu_{tjg})\\
\log(\mu_{tjg}) &= \alpha_1 I(X_{1tg} < c^{\star}) X_{1tg} + \alpha_2 I(X_{1tg}\geq c^{\star})X_{1tg} + \sum_{m=1}^4\beta_mZ_{mtg} + u_{j}\\
u_j &\sim N(0, 100)\\
\alpha_1, \alpha_2, \beta_1, \dots \beta_4 & \sim N(0, 1)\\
\end{align*}

---

# Step 2b. Spatial Effect modification
.pull-left[    
\begin{align*}
Y_{tj} & \sim \text{Poisson}(\mu_{tj})\\
\log(\mu_{tj}) & = \alpha_1 I(X_{1t}< c^{\star})X_{1t} + \alpha_{2s_j} I(X_{1t}\geq c^{\star})X_{1t} +\\ 
& \sum_{m=1}^4\beta_mZ_{mt} + u_{j}\\
\alpha_{2s} &= \alpha_2 + \sum_{m=1}^8 \gamma_q H_{sq} + v_s + b_s\\
v_s &\sim N(0, \sigma_2)\\
b_s|b_{-s} &\sim N\left(\frac{\sum_{s \sim r} w_{rs} b_s}{\sum_{s \sim r} w_{rs}},        \frac{\sigma^2}{\sum_{s \sim r} w_{rs}}\right)\\
\end{align*}

where $\bm{H}_s$ is a vector of the following covariates at the Lower Tier Local Authority level (LTLA):

1\. Quintiles of deprivation
2\. Proportion of green space
3\. Urbanicity index
4\. Average temperature
]

.pull-right[
  `r include_fig("London.jpg")`
]

`r vspace("30px")`
.small[
Konstantinoudis et al., Asthma hospitalisations and heat exposure in England: a case–crossover study during 2002–2019, Thorax (2023), in press.
]

---

# Results

.panelset[.panel[.panel-name[Threshold selection and effect modification]

- In the adjusted model, the 80th percentile of the temperature (23.2C) was the threshold minimising the WAIC

- The increase in the COPD hospitalisation risk for every 1°C increase in temperatures above 23.2C was 0.47% (95% CrI 0.16% to 0.75%) 


.pull-left[

`r vspace("30px")`
Effect modification by age and sex: 

- unadjusted models show increased risk which substantially reduces when the confounders are included
`r vspace("15px")`
- Evidence of an increased risk for females above 65 years old
]

.pull-right[
`r include_fig("ConfoundersRR.jpg", width="100%")`
]
]

.panel[.panel-name[Spatial effect modifiers]

.pull-left[
`r include_fig("SVC.png", width="100%")`

`r include_fig("SVC_adj.png", width="100%")`

]

.pull-right[

- Evidence of spatial variation
`r vspace("30px")`
- Not enough evidence of effects of the covariates in explaining the spatial variation

```{r res-tables}
library(kableExtra)
Modifier<- c("Urban","Avg Temp")
PercentChange <- c("−1.57 (−4.16 to 0.96)","−0.41 (−1.49 to 0.71)")
Probability<-c(0.12,0.22)

df<- data.frame(Modifier,PercentChange,Probability)
df %>% 
  kbl() %>% 
  kable_material(c("striped", "hover"))
```
]
]
]

---

# Population Attributable Fraction (PAF)

- Typically used in epidemiological studies to evaluate the number of health end points which happened due to the exposure/risk factor.

.pull-left[
- Generally estimated over the entire study region

- Here we used the model with spatially-varying coefficients to estimate a local PAF using the following three steps

1. Obtain the cumulative exposure-outcome relative risk $\text{RR}_s = \exp\big(\alpha_{2s}(X_{1, s, \text{max}} - c^*)\big)$
2. Calculate $\text{AF}_s = (\text{RR}_s - 1)/\text{RR}_s$

3. Then $\text{PAF}_s = \text{AF}_s (n_s/N_s)$

where $N_s$ and $n_s$ are the total number of hospitalisations and those at days above 23.2C 
]

.pull-right[

`r include_fig("PAF.jpg", width="70%")`

- Overall 7.8% (95% CrI 6.7% to 8.8%) of the total COPD hospitalisations in summer (2007-2019) are due to temperature above 23.2C
- Clear spatial structure (more than 8% in East Midlands, East of England, London and South East, while it is below 5% in the South West).

]

---

# Take home message

- Spatial case-crossover model allows to take advantage of individual level data when the individual level confounders are not available (as long as they do not change quickly in time)

`r vspace("20px")`

- Extended to investigate spatially effect modifiers (could be done for temporal effect modifiers too)

`r vspace("20px")`

- Computationally intensive

`r vspace("20px")`

- Evidence that COPD hospitalisations increase with warmer temperatures and as temperatures consistently increase, public health systems should be alerted and prepared to challenge the increased COPD hospitalisation burden. 

---
# General remarks

- Spatial and temporal dependencies are crucial to account for in environmental epidemiology

`r vspace("10px")`

$\rightarrow$ they add a level of understanding on the phenomenon under study and help reduce biases

`r vspace("10px")`

$\rightarrow$ using the spatio-temporal structure (and potentially covariates) the model can predict at any location in the study region

`r vspace("20px")`

- Increased flexibility, but also additional statistical challenges

`r vspace("10px")`

$\rightarrow$ parameter structure

`r vspace("10px")`

$\rightarrow$ computationally intensive

`r vspace("10px")`

$\rightarrow$ transferability

`r vspace("20px")`

- Reporting is challenging, particularly when dealing with stakeholders

---

# Other areas of work

`r vspace("20px")`

- Evaluating the impact of short and long term aircraft noise on cardiovascular hospitalisation and mortality

`r vspace("20px")`

- Modelling excess mortality at subnational level across different countries, to estimate the impact of extreme events (e.g. heat, COVID, policies)

`r vspace("20px")`

- Examine heat related vulnerabilities in people living with diabetes (ICRF grant)

`r vspace("20px")`

- Apportioning particulate matter into sources using Bayesian parametric and non parametric models to evaluate the effects on cardio-respiratory healths in children and elderly

---
# Acknowledgments

.red[Wastewater based epidemiology:]

**Guangquan Li** (Northumbria University)

**Radka Jersakova**, **Callum Mole**, **Camila Rangel Smith**  (Alan Turing Institute)

**Barry Rowlingson**, **Peter Diggle** (Lancaster University)

**Sylvia Richardson** (MRC Biostatistics Unit, Cambridge)

**George Nicholson**, **Chris Holmes** (Oxford University)

**Fatemeh Torabi** (Swansea University)

**Matt Wade**, **Hubert Denise**, **William Rowe**, **Daniel James**, **Jasmine Grimsley** (UKHSA)

`r vspace("20px")`

.red[Temperature and COPD:]

**Cosetta Minelli** (ICL)

**Antonio Gasparrini** (LSHTM)

**Ana Maria Vicedo Cabrera** (University of Bern)

**Joan Ballester** (ISGlobal)

---

# Links to papers and code

- Li G., Diggle P., Blangiardo M. **Integrating wastewater and randomised prevalence survey data for national COVID surveillance**, *Scientific Reports*, 14(5124), 2024; code:[https://github.com/gqlNU/wwprev](https://github.com/gqlNU/wwprev)

- Li G. et al, **A spatio-temporal framework for modelling wastewater concentration during the COVID-19 pandemic**, *Environment International*, 172(107765), 2023; code:[https://github.com/gqlNU/publicWW/](https://github.com/gqlNU/publicWW/)

